---
title: "Introduction to $\\mathtt{R}$"
subtitle: "Session 5 -- Basic Graphics"
author: "Statistical Consulting Centre"
date: "20 July, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = FALSE)
# knitr::opts_chunk$set(eval = FALSE)

lake.df <- read.csv("..\\..\\data\\Lake.csv", 
                      stringsAsFactors = FALSE)

lake.df$pHtype <- with(lake.df, ifelse(pH > 7, "alkalinity", ifelse(pH < 7, "acidity", "natural")))
lake.df$pHtype <- factor(lake.df$pHtype, levels = c( "acidity", "natural", "alkalinity"))
lake.df$Calcium <- factor(lake.df$Calcium, levels = c("Low", "Medium", "High"))

mercury.df <- read.csv("..\\..\\data\\Mercury.csv", 
                      stringsAsFactors = FALSE)
library(dplyr)
joined.df <- left_join(lake.df, mercury.df, by = "ID")

library(tidyr)
joined.long.df <- gather(joined.df, "Day1", "Day2", "Day3", key = Time, value= Mercury)

joined.long.df$Calcium <- as.character(joined.long.df$Calcium)
```


# 1. Boxplot

1. Draw boxplots showing the distribution of \texttt{Mercury} for each of the levels of \texttt{Calcium}. The boxplot should look exactly like 

```{r}
with(joined.long.df, plot(as.factor(Calcium), Mercury,
                      xlab = "Calcium level", cex.lab = 1.5,
                      ylab = "Mercury concentration",
                      main = "Mercury contaminations in Florida lakes"))

```

The boxplot should have:
\begin{itemize}
  \item labels on the x-axis,
  \item an informative label on the y-axis (not \texttt{Mercury}),
  \item a title, and
  \item labels that are large enough to read without squinting.
\end{itemize}

- Now try to produce the boxplot so that \texttt{Calcium} is plotted in a sensible order with respect to \texttt{Calcium} level like
```{r}
joined.long.df$Calcium <- factor(joined.long.df$Calcium, level = c("Low", "Medium", "High"))
with(joined.long.df, plot(Calcium, Mercury,
                      xlab = "Calcium level", cex.lab = 1.5,
                      ylab = "Mercury concentration",
                      main = "Mercury contaminations in Florida lakes"))
```

Hint: use the following command to convert \texttt{Calcium} variable for plotting.

```
joined.long.df$Calcium <- factor(joined.long.df$Calcium, level = c("Low", "Medium", "High"))
```


# 2. Scatterplot
Draw a scatter plot which shows the relationship between \texttt{pH} and \texttt{Mercury} for each of the \texttt{Calcium} levels. The scatter plot should look exactly the same as 

```{r}
with(joined.long.df, plot(pH, Mercury, 
     col = c(1,2,4)[Calcium], 
     pch = c(20, 22,3)[Calcium], 
     xlab = "pH value", ylab = "Mercury concentration"))

legend("bottomleft", pch = c(20, 22, 3), 
       col = c(1, 2, 4),
       bty = "n", cex = 1.5,
legend = c("Low", "Medium", "High"))
```

Hint: 

- \texttt{Low} \texttt{Calcium} level: using \texttt{pch = 20} and \texttt{col = 1}.
- \texttt{Medium} \texttt{Calcium} level: using \texttt{pch = 22} and \texttt{col = 2}.
- \texttt{High} \texttt{Calcium} level: using \texttt{pch = 3} and \texttt{col = 4}.


# 3. Barplot

1. Draw a juxtaposed barplot that shows the average \texttt{Mercury} concentration for the six combinations of \texttt{Calcium} and \texttt{pHtype}. The graph should look exactly the same as

```{r}
trt.mean <- with(joined.long.df, tapply(Mercury,
                            list(pHtype, Calcium), mean))
barplot(trt.mean, beside = TRUE, space = c(0.1, 1),
legend.text= c("Acidity", "Neutral", "Alkalinity"),
        ylab = "Mean mercury concentration", 
        xlab = "Calcium level",
args.legend = list(bty = "n", cex = 1.5))
```

Draw the barplot in the following steps:

- Calculate the mean \texttt{mercury} level for each combination of \texttt{pHtype} and \texttt{plot.cal}, using the \texttt{tapply()} function.
- Save the calculated means in a new object called \texttt{trt.mean}.
- Produce the barplot using \texttt{beside = TRUE}, putting a  small space  between the bars. (Reminder: the argument that does this is \texttt{space}).
- Add a legend.

2. Mess up the plot by playing with \texttt{horiz}, \texttt{angle}  \texttt{density} and any arguments that you want. Note: keep this to yourself.



