---
title: "Introduction to $\\mathtt{R}$"
header-includes: 
   \usepackage{graphicx}
    \subtitle{Session 5 -- Basic Graphics}    
    \author[SCC]{Statistical Consulting Centre}
   \institute[\href{mailto:consulting@stat.auckland.ac.nz}
    {consulting@stat.auckland.ac.nz}]{\href{mailto:consulting@stat.auckland.ac.nz}
    {consulting@stat.auckland.ac.nz}\\
  The Department of Statistics\\
  The University of Auckland}
   \titlegraphic{\includegraphics[width=5cm]{..//..//S-DS-VC-RGB.png}}
date: "20 July, 2017"
output:
  beamer_presentation:
    theme: "Madrid"
    colortheme: "crane"
    fonttheme: "professionalfonts"
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(tidy = TRUE)
knitr::opts_chunk$set(message = FALSE)

options(width = 50)
library(dplyr)
library(tidyr)

Patient.df <- read.csv("..\\..\\data\\Patient.csv", stringsAsFactors = FALSE)
Patient.df$Smoke.group <- 
  with(Patient.df, ifelse(Smoke == 1, "Yes", "No"))
table(Patient.df$Smoke.group)
Patient.df$Race.group <- 
  with(Patient.df,
       ifelse(Race == 1, "Caucasian",
              ifelse(Race == 2, "African", "Other")))
table(Patient.df$Race.group)
Patient.df$BMI <- with(Patient.df, (Weight/Height^2)*703)
Patient.df$BMI.group <- 
  ifelse(Patient.df$BMI >= 30, "obese", 
         ifelse(Patient.df$BMI >= 25, "overweight",
          "normal"))

Patient.df <- Patient.df[, !names(Patient.df) %in% c("Smoke", "Race")]

Patient.df$BMI.group <- factor(Patient.df$BMI.group,
               levels = c("normal", "overweight", "obese"))

Patient.df$Sex <- factor(Patient.df$Sex)
Patient.df$Race.group <- factor(Patient.df$Race.group)
Patient.df$Smoke.group <- factor(Patient.df$Smoke.group)

Patient.df$Age.group <- 
  ifelse(Patient.df$Age<=35, "Under 35", 
         ifelse(Patient.df$Age <= 60, "36 to 60", "Over 61"))
Patient.df$Age.group <- factor(Patient.df$Age.group, 
                           levels = c("Under 35", "36 to 60", "Over 61"))
Cholesterol.df <- read.csv("..\\..\\data\\CholesterolNA.csv")

library(dplyr)

combined.df <- left_join(Patient.df, Cholesterol.df)

names(combined.df)[c(1,11,12,13)] <- 
  c("ID", "Baseline", "PreTrt", "PostTrt")

combined.long.df <- gather(combined.df, key = Time,  value= Cholesterol, 
                           -ID, -Age, -Age.group, -Sex, -Weight, -Height,
                           -Smoke.group, -Race.group, -BMI,
                           -BMI.group)

```


# Exploring your data

- Examine for patterns, relationships, structures, and other features.
- Do this using graphs, tables and summary statistics.
- Patterns are more easily seen in graphs.
- Remember:
    \begin{center}
      ``\texttt{R} is a free software environment for statistical computing and \emph{graphics}''
    \end{center}
    
    
# Designing graphs

Two important considerations when designing a graph:

- It should have something to say.
    + Decide what information you want your graph to display.
    + Graphical software is capable of producing any number of different graphs, so generating graphs with meaningless information is quite simple to do!

- It should be easy to interpret.
    + Simplicity is key!
    + Sufficiently large fonts, all axes labelled, clearly defined legends, etc.


# Some functions for drawing graphs

- Usually, we use \texttt{plot()} to create a graph and then \texttt{lines()}, \texttt{points()}, \texttt{text()}, and other commands to annotate it.
- \texttt{plot()} does appropriate things for different types of variables (see later examples).
- Variables of type \texttt{cha}racter are not suitable for plotting (and analysis). Convert such variables to type \texttt{factor}.
- There are no "erase" or "undo" functions, so store your commands in an \texttt{R} script.


# Boxplots

```{r, fig.align='center'}
with(combined.long.df, plot(Sex, Cholesterol))
```


# Boxplots

```{r, error=TRUE}
with(combined.long.df, plot(as.character(Sex), Cholesterol))
```


# Boxplots

`Sex` varaible in `combined.long.df` has to be `factor`. 

```{r}

class(combined.long.df$Sex)

```

#Formula interface
The plot() command can also be used this way:

```{r, fig.height = 5, fig.width = 7, fig.align='center'}
with(combined.long.df, plot(Cholesterol ~ Sex))
```


# Axes need labels

Label axes with units.

```{r, eval = FALSE, tidy = FALSE}
with(combined.long.df, 
     plot(Sex, Cholesterol, xlab = "Gender",
          ylab = "Cholesterol level (mg/100ml)"))
```


# `xlab` and `ylab`

```{r, echo=FALSE, fig.align='center'}
with(combined.long.df, 
     plot(Sex, Cholesterol, xlab = "Gender",
          ylab = "Cholesterol level (mg/100ml)"))
```

# Axes labels are too small

Label axes with units, large enough to read.

```{r, eval = FALSE, tidy = FALSE}
with(combined.long.df, 
     plot(Sex, Cholesterol, xlab = "Gender",
          ylab = "Cholesterol level (mg/100ml)",
          cex.lab = 1.5))
```

# `cex.lab = 1.5`

```{r, echo=FALSE, fig.align='center'}
with(combined.long.df, 
     plot(Sex, Cholesterol, xlab = "Gender",
          ylab = "Cholesterol level (mg/100ml)",
          cex.lab = 1.5))
```

# Title

Sometimes you need a title.

```{r, eval = FALSE, tidy = FALSE}
with(combined.long.df, 
     plot(Sex, Cholesterol, xlab = "Gender",
          ylab = "Cholesterol level (mg/100ml)",
          cex.lab = 1.5, 
          main = "Cholesterol level for male and female"))
```


# `main`
```{r, echo=FALSE, fig.align='center'}
with(combined.long.df, 
     plot(Sex, Cholesterol, xlab = "Gender",
          ylab = "Cholesterol level (mg/100ml)",
          cex.lab = 1.5, 
          main = "Cholesterol level for male and female"))
```


# `boxplot()`
`boxplot()` is probably more fexible than `plot()`, if you are drawing boxplots. Check the help document for more details.

```{r, echo=FALSE, fig.align='center'}
with(combined.long.df, boxplot(Cholesterol~Sex))
```


# Smoking group by Age group

Examine the data via twp=way freqency table. 

```{r} 
with(combined.df, table(Smoke.group, Age.group))


```

# Smoking group by Age group

Then, convert the counts to percentages.

```{r, tidy = FALSE}
smoke.age <- with(combined.df, 100* 
                    prop.table(table(Smoke.group, 
                                     Age.group), 2))

round(smoke.age, 2)

```

# `barplot`
    
```{r, eval = FALSE, tidy = FALSE}
barplot(smoke.age, xlab = "Age group by years", 
        ylab = "Percentage")

```
    
# `barplot`
    
```{r, echo=FALSE, fig.align='center'}
barplot(smoke.age, xlab = "Age group by years", ylab = "Percentage")

```   
    
# Side-by-side `barplot`
    
```{r, eval = FALSE, tidy = FALSE}
barplot(smoke.age, beside = TRUE,
        xlab = "Age group by years", 
        ylab = "Percentage")

```
    
# `beside = TRUE`
    
```{r, echo=FALSE, fig.align='center'}
barplot(smoke.age, beside = TRUE, xlab = "Age group by years", ylab = "Percentage")

```      
    
# Line along horizontal axis?
    
```{r, eval = FALSE, tidy = FALSE}
barplot(smoke.age, beside = TRUE,
        xlab = "Age group by years", 
        ylab = "Percentage")
abline(h = 0)
```
    
# `abline(h = 0)`
    
```{r, echo=FALSE, fig.align='center'}
barplot(smoke.age, beside = TRUE, xlab = "Age group by years", ylab = "Percentage")
abline(h = 0)
```  

# Vertical axis higher than bars?

```{r, eval = FALSE, tidy = FALSE}
barplot(smoke.age, beside = TRUE,
        xlab = "Age group by years", 
        ylab = "Percentage", 
        ylim = c(0, 100))
abline(h = 0)
```

# `ylim = c(0, 100)`
    
```{r, echo=FALSE, fig.align='center'}
barplot(smoke.age, beside = TRUE, xlab = "Age group by years", ylab = "Percentage", ylim = c(0, 100))
abline(h = 0)
```  

# Adding space between bars 

```{r, eval = FALSE, tidy = FALSE}
barplot(smoke.age, beside = TRUE,
        xlab = "Age group by years", 
        ylab = "Percentage", 
        ylim = c(0, 100),
        space = c(0.2, 1.5))
abline(h = 0)
```

# `space = c(0.2, 1.5)`
    
```{r, echo=FALSE, fig.align='center'}
barplot(smoke.age, beside = TRUE, 
        xlab = "Age group by years", 
        ylab = "Percentage", 
        ylim = c(0, 100),
        space = c(0.2, 1.5))
abline(h = 0)
```  

# Shading with lines

```{r, eval = FALSE, tidy = FALSE}
barplot(smoke.age, beside = TRUE,
        xlab = "Age group by years", 
        ylab = "Percentage", 
        ylim = c(0, 100),
        space = c(0.2, 1.5), 
        density = c(0, 100))
abline(h = 0)
```

# `density = c(0, 100)`
    
```{r, echo=FALSE, fig.align='center'}
barplot(smoke.age, beside = TRUE, 
        xlab = "Age group by years", 
        ylab = "Percentage", 
        ylim = c(0, 100),
        space = c(0.2, 1.5), 
        density = c(0, 100))
abline(h = 0)
```  

# Shading with lines
The defualt colour for shading lines is pale while! Change to black using the `col` argument. 

```{r, echo=FALSE, fig.align='center'}
barplot(smoke.age, beside = TRUE, 
        xlab = "Age group by years", 
        ylab = "Percentage", 
        ylim = c(0, 100),
        space = c(0.2, 1.5), 
        density = c(0, 100), 
        col = "black")
abline(h = 0)
```  

# `col = "black"`
```{r, echo=FALSE, fig.align='center'}
barplot(smoke.age, beside = TRUE, 
        xlab = "Age group by years", 
        ylab = "Percentage", 
        ylim = c(0, 100),
        space = c(0.2, 1.5), 
        density = c(0, 100), 
        col = "black")
abline(h = 0)
```  

# Color-filled bars? 

```{r, eval = FALSE, tidy = FALSE}
barplot(smoke.age, beside = TRUE,
        xlab = "Age group by years", 
        ylab = "Percentage", 
        ylim = c(0, 100),
        space = c(0.2, 1.5), 
        col = c("blue", "green"))
abline(h = 0)
```

# `col = c("blue", "green")`
    
```{r, echo=FALSE, fig.align='center'}
barplot(smoke.age, beside = TRUE, 
        xlab = "Age group by years", 
        ylab = "Percentage", 
        ylim = c(0, 100),
        space = c(0.2, 1.5), 
        col = c("blue", "green"))
abline(h = 0)
```  

# Legend? 

```{r, eval = FALSE, tidy = FALSE}
barplot(smoke.age, beside = TRUE,
        xlab = "Age group by years", 
        ylab = "Percentage", 
        ylim = c(0, 100),
        space = c(0.2, 1.5), 
        legend.text = TRUE)
abline(h = 0)
```

# `legend.text = TRUE`
    
```{r, echo=FALSE, fig.align='center'}
barplot(smoke.age, beside = TRUE, 
        xlab = "Age group by years", 
        ylab = "Percentage", 
        ylim = c(0, 100),
        space = c(0.2, 1.5), 
        legend.text = TRUE)
abline(h = 0)
``` 

# Further improvements

- Larger title for axes
- Get rid of the box drawn outide the legend
- Larger legend

```{r, eval = FALSE, tidy = FALSE}
barplot(smoke.age, beside = TRUE,
        xlab = "Age group by years", 
        ylab = "Percentage", 
        ylim = c(0, 100),
        space = c(0.2, 1.5), 
        legend.text = TRUE, 
        args.legend = list(bty = "n", cex = 1.3))
abline(h = 0)
```

# ` args.legend = list(bty = "n", cex = 1.3)`
    
```{r, echo=FALSE, fig.align='center'}
barplot(smoke.age, beside = TRUE, 
        xlab = "Age group by years", 
        ylab = "Percentage", 
        ylim = c(0, 100),
        space = c(0.2, 1.5), 
        legend.text = TRUE, 
        args.legend = list(bty = "n", cex = 1.3))
abline(h = 0)
``` 

# Histgrams 
```{r, tidy = FALSE, fig.height = 5, fig.width = 7, fig.align='center'}
hist(combined.long.df$Cholesterol)
```

# Fewer bins?  
```{r, tidy = FALSE, fig.height = 5, fig.width = 7, fig.align='center'}
hist(combined.long.df$Cholesterol, breaks = 10)
```

# More bins?  
```{r, tidy = FALSE, fig.height = 5, fig.width = 7, fig.align='center'}
hist(combined.long.df$Cholesterol, breaks = 40)
```

# Always use meaningful axis labels
```{r, tidy = FALSE, fig.height = 3, fig.width = 6, fig.align='center', tidy = FALSE}
hist(combined.long.df$Cholesterol, main = NULL, 
     xlab = "Cholesterol level (mg/100ml)")
```

# Scatter plots

- Used to dsplay paired quantiative (numeric) data, e.g. age and Cholesterol level. 
- Make sure they are numeirc. 
- Use `plot()` function. 

```{r, eval = FALSE, tidy = FALSE}
with(combined.long.df, plot(Age, Cholesterol))

```

# Scatter plots

```{r, echo=FALSE, fig.align='center'}
with(combined.long.df, plot(Age, Cholesterol))

```

# `type =` 

type = controls how data are plotted.

- `type = "p"`: points, this is the default,
- `type = "l"`: lines,
- `type = "b"`: both lines and points,
- `type = "n"`: no plotting, which can set up a plot for latter additions.

# Plotting character

We will usually use different plotting character for different groups.
Argument pch = in plot() or points() is used to change the plotting character:

- `pch = 19`: solid circle,
- `pch = 20`: smaller solid circle,
- `pch = 21`: filled circle,
- `pch = 22`: filled square,
- `pch = 23`: filled diamond,
- `pch = 24`: filled triangle point up,
- `pch = 25`: filled triangle point down.

# Scatter plots

```{r, eval = FALSE, tidy = FALSE}
with(combined.long.df, plot(Age, Cholesterol, 
     col = c("black", "red")[Sex], 
     pch = c(19, 24)[Smoke.group]))
legend("topright", pch = c(19, 19, 24, 24),
      col = c("black", "red", "black", "red"),
      pt.bg = c("black", "red", NA, NA),
      legend = c("Non-Smokers:Female", "Non-Smokers:Male",
                "Smokers:Female", "Smokers:Male"))
```

# Scatter plots

```{r, echo=FALSE, fig.align='center'}
with(combined.long.df, plot(Age, Cholesterol, 
     col = c("black", "red")[Sex], 
     pch = c(19, 24)[Smoke.group]))
legend("topright", pch = c(19, 19, 24, 24),
      col = c("black", "red", "black", "red"),
      pt.bg = c("black", "red", NA, NA),
      legend = c("Non-Smokers:Female", "Non-Smokers:Male",
                "Smokers:Female", "Smokers:Male"))
```


# Graphical Parameters

```{r, eval = FALSE, tidy = FALSE}
?par
op <- par(mfrow = c(2, 2))       
  with(combined.long.df, boxplot(Cholesterol~Age.group,
      xlab = "Age Group", ylab = "Total score"))
  barplot(smoke.age, beside = TRUE, ylab = "Percentage",
      xlab = "Age group in years", ylim = c(0, 100),
      space = c(0.2, 1.5), legend.text = TRUE)
      abline(h = 0)
  hist(combined.long.df$Age)
  with(combined.long.df, plot(Age, Cholesterol))
## At end of plotting, reset to previous settings:
par(op)

```

# Graphical Parameters

```{r, echo=FALSE, fig.align='center'}
op <- par(mfrow = c(2, 2))       
  with(combined.long.df, boxplot(Cholesterol~Age.group,
      xlab = "Age Group", ylab = "Total score"))
  barplot(smoke.age, beside = TRUE, ylab = "Percentage",
      xlab = "Age group in years", ylim = c(0, 100),
      space = c(0.2, 1.5), legend.text = TRUE)
      abline(h = 0)
  hist(combined.long.df$Age)
  with(combined.long.df, plot(Age, Cholesterol))
## At end of plotting, reset to previous settings:
par(op)

```

# Save your graph















































    
