---
title: "Introduction to $\\mathtt{R}$"
subtitle: "Session 1 -- Introduction"
author: "Statistical Consulting Centre"
date: "19 July, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(eval = FALSE)
```

# 1. Using R as a calculator

1. Find the values of:

    (a) $1+4$

    ```{r}
      1+4
    ```
    
    (b) $2^3 + \frac{4}{\sqrt{34}}$
    
    ```{r}
      2^3 + 4/sqrt(34)
    ```
    
    (c) $\log{30}$
    
    ```{r}
      log(30)
    ```
    
    (d) $\log_{10}30$
    
    ```{r}
      log(30)
    ```
    
    (e) $|-2|$ \hspace{0.2cm}(Hint: $|x|$ denotes the \emph{absolute
    value} of $x$. Search on Google if you're unsure.)
    
    ```{r}
      abs(-2)
    ```

2. Now open Rstudio, open a \texttt{R} script clicking \texttt{File}
  $\rightarrow$ \texttt{New} $\rightarrow$ \texttt{R} script.
  
3. Save this script by clicking \texttt{File} $\rightarrow$
  \texttt{Save As...}.
  
4. Select a directory/location and save the script. Note: the saved
  script should have \texttt{.r} as extension. For example, if you
  call your file \texttt{exercise one}, then you should save it as
  \texttt{exercise one.r}
  
5. Copy and paste the code you typed (\emph{not the output, not the
  $>$ symbol, just the code you typed}) at the console for into the \texttt{R} script opened in Rstudio.
  
6. Submit your entire script at once to the \texttt{R}
  Console by highlighting all codes and pressing Ctrl $+$ R.
  
7. From now on, type all of your code in your \texttt{R} script and
  submit it to the \texttt{R} Console using Ctrl $+$ R.
  
  
# 2. Reading data into \texttt{R}

1. \texttt{lake.csv} contains data on mercury contamination in
  53 different lakes in Florida. The variable names and what has been
  measured are presented below.
  \begin{itemize}
  \item ID: ID number of the lake
  \item Lake: Name of the lake
  \item pH: pH value
  \item Calcium: concentration of Calcium
  \item Chlorophyll: concentration of Chlorophyll (mg/L)
  \item Avg\_Mercury: Average mercury concentration (parts per
    million) in the muscle tissue of the fish sampled from that lake.
  \end{itemize}
  
2. Read the data into \texttt{R}, saving it in object named \texttt{mercury.df}.

3. Use \texttt{dim()} and \texttt{head()} to look at some of the  properties of the dataset you have just read into \texttt{R}. \emph{Always} perform this important step to check that your dataset is as it should be.

4. Calculate the mean and standard deviation of both pH and
  Chlorophyll.

5. Check out what \texttt{summary()} does by running  \texttt{summary(mercury.df\$pH)}.





