---
title: "Introduction to $\\mathtt{R}$"
subtitle: "Session 1 -- Introduction"
author: "Statistical Consulting Centre"
date: "19 July, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = FALSE)
# knitr::opts_chunk$set(eval = FALSE)
```

# 1. Using `R` as a calculator

1. Find the values of:

    (a) $1+4$

    ```{r}
      1+4
    ```
    
    (b) $2^3 + \frac{4}{\sqrt{34}}$
    
    ```{r}
      2^3 + 4/sqrt(34)
    ```
    
    (c) $\log{30}$
    
    ```{r}
      log(30)
    ```
    
    (d) $\log_{10}30$
    
    ```{r}
      log(30)
    ```
    
    (e) $|-2|$ \hspace{0.2cm}(Hint: $|x|$ denotes the \emph{absolute
    value} of $x$. Search on Google if you're unsure.)
    
    ```{r}
      abs(-2)
    ```

2. Now open Rstudio, open a \texttt{R} script clicking \texttt{File}
  $\rightarrow$ \texttt{New} $\rightarrow$ \texttt{R} script.
  
3. Save this script by clicking \texttt{File} $\rightarrow$
  \texttt{Save As...}.
  
4. Select a directory/location and save the script. Note: the saved
  script should have \texttt{.r} as extension. For example, if you
  call your file \texttt{exercise one}, then you should save it as
  \texttt{exercise one.r}
  
5. Copy and paste the code you typed (\emph{not the output, not the
  $>$ symbol, just the code you typed}) at the console for into the \texttt{R} script opened in Rstudio.
  
6. Submit your entire script at once to the \texttt{R}
  Console by highlighting all codes and pressing Ctrl $+$ R.
  
7. From now on, type all of your code in your \texttt{R} script and
  submit it to the \texttt{R} Console using Ctrl $+$ R.
  
  
# 2. Reading data into \texttt{R}

1. \texttt{Lake.csv} contains data on 53 different lakes in Florida. The variable names and what has been measured are presented below.
    + `ID`: ID number of the lake
    + `Lake`: Name of the lake
    + `pH`: pH value
    + `Calcium`: concentration of Calcium
    + `Chlorophyll`: concentration of Chlorophyll (mg/L)
    

2. Read the data into \texttt{R}, saving it in object named \texttt{lake.df}.

```{r, eval = FALSE}
lake.df <- read.csv("location of your folder/Lake.csv", 
                      stringsAsFactors = FALSE)

```

```{r, echo = FALSE}
lake.df <- read.csv("..\\..\\data\\Lake.csv", 
                      stringsAsFactors = FALSE)
```

3. Use \texttt{dim()} and \texttt{head()} to look at some of the  properties of the dataset you have just read into \texttt{R}. \emph{Always} perform this important step to check that your dataset is as it should be.

```{r}
dim(lake.df)
head(lake.df)
```

4. Calculate the mean and standard deviation of both `pH` and `Chlorophyll`.

```{r}
mean(lake.df$pH, na.rm = TRUE)
mean(lake.df$Chlorophyll, na.rm = TRUE)

sd(lake.df$pH, na.rm = TRUE)
sd(lake.df$Chlorophyll, na.rm = TRUE)

```

5. Check out what \texttt{summary()} does by running  \texttt{summary(lake.df\$pH)}.

```{r}
summary(lake.df$pH)

```

6. Check the ferquency of `Calcium` concentration

```{r}
table(lake.df$Calcium)

```

7. Turn the freuqency table in 2.6 into proportion, keep only 2 decimal places.

```{r}
round(prop.table(table(lake.df$Calcium)) * 100, 1)

```