---
title: "Introduction to $\\mathtt{R}$"
header-includes: 
   \usepackage{graphicx}
    \subtitle{Session 2 -- Data manipulation}    
    \author[SCC]{Statistical Consulting Centre}
   \institute[\href{mailto:consulting@stat.auckland.ac.nz}
    {consulting@stat.auckland.ac.nz}]{\href{mailto:consulting@stat.auckland.ac.nz}
    {consulting@stat.auckland.ac.nz}\\
  The Department of Statistics\\
  The University of Auckland}
   \titlegraphic{\includegraphics[width=5cm]{..//..//S-DS-VC-RGB.png}}
date: "19 July, 2017"
output:
  beamer_presentation:
    theme: "Madrid"
    colortheme: "crane"
    fonttheme: "professionalfonts"
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(tidy = TRUE)
options(width = 50)
Patient.df <- read.csv("..\\..\\data\\Patient.csv", stringsAsFactors = FALSE)
```

#Functions
A function is a relationship between a set of inputs (arguments) and a set of outputs. E.g., the function is fed some information on which it operates, the results of which are the output.

\vspace{-30pt}

```{r, echo = FALSE, message=FALSE, fig.asp=0.5}    
require(plotrix)
plot.new()
plot.window(xlim = c(-8, 8), ylim = c(-1, 1))
draw.circle(-6, 0, 2)
draw.circle(0, 0, 2)
draw.circle(6, 0, 2)
text(c(-6, 0, 6), 0, c("Input", "Function", "Output"), cex = 1.4)
arrows(c(-3.75, 2.25), 0, c(-2.25, 3.75), 0, length = 0.1)
```
\vspace{-50pt}

This is an essential building block for the \texttt{R} package.    


# Functions
 We have seen many functions, e.g. \texttt{log}, \texttt{mean}, \texttt{table}, \texttt{with}, etc.

 \vspace{-30pt}
```{r, echo = FALSE, message=FALSE, fig.asp=0.5}    
plot.new()
plot.window(xlim = c(-8, 8), ylim = c(-1, 1))
draw.circle(-6, 0, 2)
draw.circle(0, 0, 2)
draw.circle(6, 0, 2)
text(c(-6, 0, 6), 0, c("1, 2, 3, 4", "Mean", "2.5"), cex = 1.4)
arrows(c(-3.75, 2.25), 0, c(-2.25, 3.75), 0, length = 0.1)
```


# Working with functions

- Functions can be user-defined, i.e., you can write your own.
- Output is the last line of the function. You can use \texttt{return()} to specifiy the output.
- Here is a function calculates the standard error of the mean (SEM).

```{r}
mystder <- function(x){
    mysd <- sd(x, na.rm = TRUE) # Calc std. deviation
    n <- length(x)              # Calc sample size
    mysd/sqrt(n)                # Definition of SEM
}
mystder(Patient.df$Height)
```

- A set of user-defined functions can be bundled together into an \texttt{R package}.

# Getting data into \texttt{R}

- Base \texttt{R} includes only functions which read data sets saved in simple file formats, e.g. \texttt{csv}, \texttt{txt}, tab delimited, etc.
- What if your data was saved in another format, e.g. Excel? 
- The \texttt{readxl} package for \texttt{R} contains functions that may help, [https://cran.r-project.org/web/packages/readxl/index.html][readxl]
  
```{r, eval = FALSE}
library(readxl)
excel <- read_excel("data.xlsx", sheet = 1)
```

[readxl]: https://cran.r-project.org/web/packages/readxl/index.html

# Getting data into \texttt{R}

- What if your data was saved in another format, e.g. Excel, STATA, SPSS, or SAS spreadsheets? 
- The \texttt{haven} package for \texttt{R} contains functions that may help, [https://cran.r-project.org/web/packages/haven/index.html][haven]
  
```{r, eval = FALSE}
library(haven)
stata <- read_dta("data.dta")
spss <- read_sav("data.sav")
sas <- read_sas("data.sas7bdat")
sasxport <- read_xpt("data.xpt")
```
However, it is always the easiest and safest to read data into \texttt{R} from a \texttt{csv} file.

[haven]: https://cran.r-project.org/web/packages/haven/index.html


# Packages
- Currently, the CRAN package repository features 10,098 available packages (4 Jul. 2017). There are about 13,169 CRAN, BioConductor and Github packages in total.
- To install packages from the \texttt{R} GUI, click on \texttt{Packages}
  $\rightarrow$ \texttt{Install Package(s) ...} 
  $\rightarrow$ \texttt{New Zealand (or whatever region you are located)} 
  $\rightarrow$ \texttt{Package name}
- Or, you can type \texttt{install.packages(\textit{package name})}, e.g. `install.packages("haven")`. 
- After the installation, use \texttt{library("\textit{package name}")} to load it into \texttt{R}. 

Note: Installation is performed only once; however, it must be loaded (i.e. use the command \texttt{library("\textit{package name}")}) in *every* \texttt{R} session.

#Bioconductor Packages

- To install core packages, type the following in an R command window:
```{r, eval = FALSE}
source("https://bioconductor.org/biocLite.R")
biocLite()
```

- Install specific packages, e.g., “GenomicFeatures” and “AnnotationDbi”, with
```{r, eval = FALSE}
biocLite(c("GenomicFeatures", "AnnotationDbi"))
```

Note: Installation is performed only once; however, it must be loaded (i.e. use the command \texttt{library("\textit{package name}")}) in *every* \texttt{R} session.


# \texttt{which} individual does Smoke?
Let's use \texttt{R}'s powerful subsetting capabilities to select those cases for which the value of Smoke is 1 (i.e. Yes).
```{r}
index <- which(Patient.df$Smoke == 1)
index
```

#How many are there?
```{r}
# Use length() to count the number of elements in "index".
length(index)
```

- `Patient.df$Smoke == 1` gives a vector of True/False for every observations.
- `which()` gives a vector of the position of `TRUE`. 
- `length()` tells us how many elements there are in the vector.

#How many are there?
Another way is to sum up how many `TRUE`s in the logical vector `Patient.df$Smoke == 1`
```{r}
sum(Patient.df$Smoke == 1)
```

#How many are there?
Another way is to sum up how many `TRUE`s in the logical vector `Patient.df$Smoke == 1`
```{r}
sum(Patient.df$Smoke == 1, na.rm = TRUE)
```

# Who are they?
```{r}
# Use square brackets to extract IDs corresponding  
# to the cases numbers contained in "index"
index <- which(Patient.df$Smoke == 1)
Patient.df$Patient.ID[index]
```

# Subsetting
Square brackets \texttt{[]} are used to extract subsets of data.
```{r}
# First element only
Patient.df$Height[1]
# All but the first element
Patient.df$Height[-1]
```

# Subsetting
Square brackets \texttt{[]} are used to extract subsets of data.
```{r}
# Elements 3 through 8
Patient.df$Height[3:8]
# Elements 3 and 8
Patient.df$Height[c(3,8)]
```

#More on subsetting
Subsetting two-dimensional arrays, such as data frames, requires the use of \em{two} indices.
```{r}
#First row or record
Patient.df[1, ] 
```

#More on subsetting
For data frames, you need to use two indices.
```{r}
#Second column or variable
Patient.df[, 2] 
```

#More on subsetting
For data frames, you need to use two indices.
```{r}
#Some rows and columns
Patient.df[2:5, 4:7] 
```

#More on subsetting
For data frames, you need to use two indices.
```{r}
#Rows by number, columns by name
Patient.df[1:10, c("Race", "Smoke")] 
```


# Missing values

-\texttt{R} reserves the object \texttt{NA} (\texttt{N}ot \texttt{A}vailable) for elements of a vector that are missing or unavailable. 
- Use of \texttt{is.na()} to search for missing values requires that they are recorded as \texttt{NA}. 
- \texttt{na} will not do because \texttt{R} is case sensitive!

```{r} 
sum(is.na(Patient.df$Smoke))
```

# Missing values
The default option of \texttt{table()} ignores \texttt{NA}s when constructing frequency tables. Now that all occurrences of \texttt{"NA, refused"} have been replaced with \texttt{NA}, missing values will no longer be shown in frequency tables constructed using \texttt{table()}.
```{r}
table(Patient.df$Smoke)
```


# Missing values
If you still want to see how many \texttt{NA}s in the frequency tables, you can change the \texttt{useNA} arguement to \texttt{"always"} in \texttt{table()}.
```{r}
table(Patient.df$Smoke, useNA = "always")
```

# Data Cleaning
The \texttt{ifelse()} function provides a quick way to convert the `Smoke` variable in `Patient.df` from number to words, i.e.
```{r, eval = FALSE}
  ifelse(test, yes, no)
```

\begin{itemize}
  \item \texttt{test}: a logical test.
  \item \texttt{yes}, what happens if the test is True.
  \item \texttt{no}, what happens if the test is False.
\end{itemize}

# Data Cleaning
```{r, tidy = FALSE}
Patient.df$Smoke.group <- 
  with(Patient.df, ifelse(Smoke == 1, "Yes", "No"))
table(Patient.df$Smoke.group)
```

# Data Cleaning
How about `Race`?
One `ifelse()` inside another `ifelse()`:
```{r, tidy = FALSE}
Patient.df$Race.group <- 
  with(Patient.df,
       ifelse(Race == 1, "Caucasian",
              ifelse(Race == 2, "African", "Other")))
table(Patient.df$Race.group)
```

# Data Cleaning
```{r}
head(Patient.df)
```

# Data Cleaning
```{r}
str(Patient.df)
```


# Your turn

First to calculate \texttt{BMI}:
\[
\mathtt{BMI} = \frac{\mathtt{Weight(kg)}}{\mathtt{Height(m)}^2}
= \frac{\mathtt{Weight(pounds)}}{\mathtt{Height(inches)}^2} \times 703
\]

Then, create a `bmi.group` for `normal`, `overweight` and `obese`, where, 

- overweight is a BMI greater than or equal to 25; and
- obesity is a BMI greater than or equal to 30.



# Your turn
To calculate \texttt{BMI}:
\[
\mathtt{BMI} = \frac{\mathtt{Weight(kg)}}{\mathtt{Height(m)}^2}
= \frac{\mathtt{Weight(pounds)}}{\mathtt{Height(inches)}^2} \times 703
\]
```{r, tidy=FALSE}
Patient.df$bmi <- (Patient.df$Weight/Patient.df$Height^2)*703
mean(Patient.df$bmi, na.rm = TRUE)
```

Or, 
```{r, tidy=FALSE}
# Calculate bmi using Weight and Height
Patient.df$bmi <- with(Patient.df, (Weight/Height^2)*703)
mean(Patient.df$bmi, na.rm = TRUE)
```

# Your turn

```{r, tidy=FALSE}
Patient.df$bmi.group <- 
  ifelse(Patient.df$bmi >= 30, "obese", 
         ifelse(Patient.df$bmi >= 25, "overweight",
          "normal"))
table(Patient.df$bmi.group)
table(Patient.df$bmi.group, useNA = "always")

```


# Subsetting in calculations

```{r}
with(Patient.df, table(Race.group, bmi.group))
```

# Subsetting in calculations
Produce the last table with Race groups of `African` and `Caucasian`, and BMI groups of `obese` and `overweight` 
```{r}
exclude.rows <-
  with(Patient.df,
     which(Race.group == "Other"  |
           bmi.group == "Normal"))

head(Patient.df[-exclude.rows, c("Race.group", "bmi.group")])
```

# Subsetting in calculations

```{r, tidy=FALSE}
R.B.tab <- with(Patient.df[-exclude.rows,], 
                table(Race.group, bmi.group))
R.B.tab
```

# Subsetting in calculations
Convert counts to percentages rounded to 1 decimal place.
```{r}
round(prop.table(R.B.tab)*100, 1)
```



# Easier way
```{r}
R.B.tab1 <- with(Patient.df, 
                 table(Race.group, bmi.group))
dim(R.B.tab1)
row.names(R.B.tab1)
colnames(R.B.tab1)
```

# Easier way
```{r}
colnames(R.B.tab1)
which(row.names(R.B.tab1) == "Other")
which(colnames(R.B.tab1) == "Normal")
```

# Easier way
```{r}
R.B.tab1[-3, -1]

round(prop.table(R.B.tab1[-3, -1])*100, 1)
```


# Subsetting in calculations
```{r, tidy = FALSE}
# Mean BMI level of females
with(Patient.df, mean(bmi[Sex == "Female"], na.rm = TRUE))

# Mean BMI level of females and Caucasian
with(Patient.df,
     mean(bmi[Sex == "Male" & 
                   Race.group == "Caucasian"],
          na.rm = TRUE))
```


# Summary

- Making R functions
- Installing and loading R pacakges
- Subsetting vectors and datasets
- `ifelse()` function











