---
title: "Introduction to $\\mathtt{R}$"
subtitle: "Session 5 -- Advanced Graphics"
author: "Statistical Consulting Centre"
date: "20 July, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = FALSE)
# knitr::opts_chunk$set(eval = FALSE)

lake.df <- read.csv("..\\..\\data\\Lake.csv", 
                      stringsAsFactors = FALSE)

lake.df$pHtype <- with(lake.df, ifelse(pH > 7, "alkalinity", ifelse(pH < 7, "acidity", "natural")))
lake.df$pHtype <- factor(lake.df$pHtype, levels = c( "acidity", "natural", "alkalinity"))
lake.df$Calcium <- factor(lake.df$Calcium, levels = c("Low", "Medium", "High"))

mercury.df <- read.csv("..\\..\\data\\Mercury.csv", 
                      stringsAsFactors = FALSE)
library(dplyr)
joined.df <- left_join(lake.df, mercury.df, by = "ID")

library(tidyr)
joined.long.df <- gather(joined.df, "Day1", "Day2", "Day3", key = Time, value= Mercury)

library(ggplot2)

```

# 1. Boxplot

1. Draw boxplots showing the distribution of \texttt{Mercury} for each of the levels of \texttt{Calcium}. The boxplot should look exactly like 

```{r}

ggplot(joined.long.df, aes(Calcium, Mercury)) + 
  geom_boxplot()+
  labs(title ="Mercury contaminations in Florida lakes",  
        x = "Calcium level", y = "Mercury concentration") +
  theme_bw()  +
  theme(plot.title = element_text(hjust = 0.5))

```

# 2. Scatterplot
Draw a scatter plot which shows the relationship between \texttt{pH} and \texttt{Mercury} for each of the \texttt{Calcium} levels. The scatter plot should look exactly the same as 

```{r}

ggplot(joined.long.df, aes(x = pH, y = Mercury, 
                           col = Calcium, shape = Calcium)) + 
  geom_point() +
  labs(title ="Mercury contaminations in Florida lakes",  
        x = "pH value", y = "Mercury concentration") +
  theme_bw()  +
  theme(plot.title = element_text(hjust = 0.5))

```

# 3. Barplot

Draw a juxtaposed barplot that shows the average \texttt{Mercury} concentration for the six combinations of \texttt{Calcium} and \texttt{pHtype}. The graph should look exactly the same as

```{r}

joined.long.df %>% 
  group_by(Calcium, pHtype) %>% 
  summarise(Mean = mean(Mercury)) %>% 
  ggplot(aes(x = Calcium, y = Mean, 
                           fill = pHtype)) + 
  geom_col(position = "dodge")  +
  labs(x = "Calcium level", y = "Mean Mercury concentration")

```



# 4. Heatmap 

Draw a heatmap that shows the \texttt{Mercury} concentration  for the six combinations of \texttt{Calcium} and \texttt{pHtype}. The graph should look exactly the same as
```{r}

ggplot(joined.long.df, aes(x = pHtype, y = Calcium)) +  +
  labs(x = "Calcium level", y = "Mean Mercury concentration")
  theme_bw()  +
  labs(y = "Calcium level", y = "pH level")

```

Hint: `geom_tile()` for the heatmap and `scale_fill_gradient` for the colour 



