---
title: "Introduction to $\\mathtt{R}$"
subtitle: "Session 5 -- Advanced Graphics"
author: "Statistical Consulting Centre"
date: "20 July, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = FALSE)
# knitr::opts_chunk$set(eval = FALSE)

lake.df <- read.csv("..\\..\\data\\Lake.csv", 
                      stringsAsFactors = FALSE)

lake.df$pHtype <- with(lake.df, ifelse(pH > 7, "alkalinity", ifelse(pH < 7, "acidity", "natural")))
lake.df$pHtype <- factor(lake.df$pHtype, levels = c( "acidity", "natural", "alkalinity"))
lake.df$Calcium <- factor(lake.df$Calcium, levels = c("Low", "Medium", "High"))

mercury.df <- read.csv("..\\..\\data\\Mercury.csv", 
                      stringsAsFactors = FALSE)
library(dplyr)
joined.df <- left_join(lake.df, mercury.df, by = "ID")

library(tidyr)
joined.long.df <- gather(joined.df, "Day1", "Day2", "Day3", key = Time, value= Mercury)

library(ggplot2)

```


# 1. Logistic regression

1. Suppose the only information available about mercury concentration were that the lakes were either clean (0) or
  contaminated (1). Create a factor called \texttt{cont}, such that \texttt{cont} $= 1$ if  \texttt{mercury} $>$ 0.53 and \texttt{cont} $= 0$ if \texttt{mercury} $\le$ 0.53.
```{r}
cont <- ifelse(joined.long.df$Mercury > 0.53, 1, 0)
cont <- as.factor(cont)
```

2. Check the one-way frequency table of \texttt{cont}.
```{r}
table(cont)
```

3. Add \texttt{cont} to \texttt{joined.long.df}.
```{r}
joined.long.df$cont <- cont
```

4. (a) Perform a logistic regression to assess how \texttt{pH} influences
  the probability of contamination.
```{r}
myglm <- glm(cont~pH, family = binomial, data = joined.long.df)
```
    (b) What is the overall significance of \texttt{pH}.
```{r}
anova(myglm, test = "Chisq")
```
    (c) How does a 1 unit increase in \texttt{pH} change the log odds of contamination?
```{r}
summary(myglm)
```

```
A 1 unit increase in pH decreases the log odds of contamination by 1.03
```
5. (a) Perform another logistic regression to check whether the probability of contamination depends on \texttt{Calcium} level.

```{r}
myglm1 <- glm(cont~Calcium, data = joined.long.df,
              family = binomial)
```

  (b) Check the overall significance of \texttt{Calcium}.
```{r}
anova(myglm1, test = "Chisq")

summary(myglm1)
```

  (c) Calculate the estimated probability of contamination for all three \texttt{Calcium} levels using \texttt{R} as calculator.
```{r}
 exp(0.8650)/(1 + exp(0.8650))

 exp(0.8650-1.4040 )/(1 + exp(0.8650-1.4040 ))
 
  exp(0.8650-2.3313)/(1 + exp(0.8650-2.3313))
```
  
  (d) Check your answer using \texttt{predict()}.
```{r}
newdata = data.frame(Calcium = c("High", "Medium", "Low"))
predict(myglm1, newdata, type = "response")
```

# 2. R Markdown








