---
title: "Introduction to $\\mathtt{R}$"
header-includes: 
   \usepackage{graphicx}
    \subtitle{Session 4 -- Data exploration}    
    \author[SCC]{Statistical Consulting Centre}
   \institute[\href{mailto:consulting@stat.auckland.ac.nz}
    {consulting@stat.auckland.ac.nz}]{\href{mailto:consulting@stat.auckland.ac.nz}
    {consulting@stat.auckland.ac.nz}\\
  The Department of Statistics\\
  The University of Auckland}
   \titlegraphic{\includegraphics[width=5cm]{..//..//S-DS-VC-RGB.png}}
date: "19 July, 2017"
output:
  beamer_presentation:
    theme: "Madrid"
    colortheme: "crane"
    fonttheme: "professionalfonts"
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(tidy = TRUE)
knitr::opts_chunk$set(message = FALSE)
options(width = 50)
Patient.df <- read.csv("..\\..\\data\\Patient.csv", stringsAsFactors = FALSE)
Patient.df$Smoke.group <- 
  with(Patient.df, ifelse(Smoke == 1, "Yes", "No"))
table(Patient.df$Smoke.group)
Patient.df$Race.group <- 
  with(Patient.df,
       ifelse(Race == 1, "Caucasian",
              ifelse(Race == 2, "African", "Other")))
table(Patient.df$Race.group)
Patient.df$BMI <- with(Patient.df, (Weight/Height^2)*703)
Patient.df$BMI.group <- 
  ifelse(Patient.df$BMI >= 30, "obese", 
         ifelse(Patient.df$BMI >= 25, "overweight",
          "normal"))

Patient.df <- Patient.df[, !names(Patient.df) %in% c("Smoke", "Race")]

Patient.df$BMI.group <- factor(Patient.df$BMI.group,
               levels = c("normal", "overweight", "obese"))

Patient.df$Sex <- factor(Patient.df$Sex)
Patient.df$Race.group <- factor(Patient.df$Race.group)
Patient.df$Smoke.group <- factor(Patient.df$Smoke.group)

Patient.df$Age.group <- 
  ifelse(Patient.df$Age<=35, "Under 35", 
         ifelse(Patient.df$Age <= 60, "36 to 60", "Over 61"))
Patient.df$Age.group <- factor(Patient.df$Age.group, 
                           levels = c("Under 35", "36 to 60", "Over 61"))
Cholesterol.df <- read.csv("..\\..\\data\\CholesterolNA.csv")

library(dplyr)

combined.df <- left_join(Patient.df, Cholesterol.df)

names(combined.df)[c(1,11,12,13)] <- 
  c("ID", "Baseline", "PreTrt", "PostTrt")

```

# `for` loop to get column summary statistics
```{r}
for (i in c("Age.group", "Sex", "Smoke.group", "Race.group", "BMI.group")){
  print(i)
  print(table(combined.df[,i]))
}

for (i in c("Age", "Height", "Weight", "BMI", "Baseline", "PreTrt", "PostTrt")){
  print(i)
  print(mean(combined.df[,i], na.rm = TRUE))
}
```

# `for` loop to get column summary statistics
```{r}

Categorical.df <- combined.df[,c("Age.group", "Sex", "Smoke.group", "Race.group", "BMI.group")]

for (i in 1:ncol(Categorical.df)){
  print(table(Categorical.df[,i]))
}

```

# `for` loop to get column summary statistics
```{r}

Continous.df <- combined.df[,c("Age", "Height", "Weight", "BMI", "Baseline", "PreTrt", "PostTrt")]


for (i in 1:ncol(Continous.df)){
  print(i)
  print(mean(Continous.df[,i], na.rm = TRUE))
}
```


# Smart way: \texttt(apply) to get column summary statistics
```{r, eval = FALSE}
apply(X, MARGIN, FUN, ...)
```
- `X`:     A data frame.
- `MARGIN`: 1 indicates rows, 2 indicates columns.
- `FUN`: function, what do you want \texttt{R} to do with the rows or columns of the data frame
- `...`: optional arguments to \texttt{FUN}.

Translation: Do something (\texttt{FUN}) to every row (or column) (\texttt{MARGIN}) of a data
frame (\texttt{X}). 


# Smart way: \texttt(apply) to get column summary statistics
```{r}
apply(Categorical.df, 2, table)
``` 


# Smart way: \texttt(apply) to get column summary statistics
```{r}
apply(Continous.df, 2, mean, na.rm = TRUE)
``` 

# Smart way: \texttt(apply) to get column summary statistics
```{r}
apply(Continous.df, 2, sd, na.rm = TRUE)
``` 

# \texttt{apply()} using self-defined \texttt{R} function
Functions used in \texttt{apply()} can be self-defined.
```{r}
na.check <- function(someinput){
 test.na <- is.na(someinput) 
 sum(test.na)
}
```
Take an educated guess at what \texttt{na.check()} does.

# \texttt{apply()} using a self-defined \texttt{R} function
Let's look at what each row of \texttt{na.check()} does.
```{r}
test1 <- Continous.df$BMI[1:10]
test1 

test.na <- is.na(test1)
test.na

sum(test.na)
```


#\texttt{apply()} using a self-defined \texttt{R} function
Let's now use \texttt{na.check()} in \texttt{apply()}.
```{r}
apply(Continous.df, 2, na.check)
```
Another way,
```{r}
apply(Continous.df, 2, function(x) sum(is.na(x)))
```


# A slightly more complicated function
```{r, tidy = FALSE}
mytab <- function(someinput){
 n <- length(someinput)
 n.missing <- na.check(someinput)
 n.complete <- n - n.missing
 mymean <- round(mean(someinput, na.rm = TRUE), 2)
 mysd <- round(sd(someinput, na.rm = TRUE), 2)
 mystder <- round(mysd/sqrt(n.complete), 2)
 Lower.CI <- round(mymean - 1.96*mystder, 2)
 Upper.CI <- round(mymean + 1.96*mystder, 2)
 c(Complete.obs = n.complete, Missing.obs = n.missing, 
   Mean = mymean, Std.Error = mystder, 
   Lower.CI = Lower.CI, Upper.CI = Upper.CI)
}
```
Take a \emph{more} educated guess at what \texttt{mytab()} does?

# A slightly more complicated function

- For the \texttt{R} novice, \texttt{mytab()} is possibly terrifying!
- We too were \texttt{R} novices once!
-  Our advice on understanding what an \texttt{R} function does?
 
  \begin{center}
  ``Use a data set for input into the function and work through it one line of code at a time.''
  \end{center}
  
- We ``experts'' still do this!
  
# mytab()

```{r}
apply(Continous.df, 2, mytab)  
```

# More descriptive stats
Calculate the mean `Baseline` score for male and female patients
```{r, tidy = FALSE}
with(combined.df, mean(Baseline[Sex == "Male"], 
                   na.rm = TRUE))
with(combined.df, mean(Baseline[Sex == "Female"], 
                   na.rm = TRUE))
```

# Better way 
```{r}
with(combined.df, tapply(Baseline, Sex, mean, na.rm = T))

```

# `tapply()`
```{r, eval=FALSE}
with(issp.df, tapply(total.lik, Income, mean, na.rm = TRUE))
```
```{r, eval=FALSE}
tapply(X, INDEX, FUN, ...)
```
Translation: Apply function \texttt{FUN} to \texttt{X} for each level in the grouping factor \texttt{INDEX}
 
How about calculating the mean `Baseline` score for each Race group (3 levels)?

# `tapply()`
```{r}
with(combined.df, tapply(Baseline, Race.group, mean, na.rm = TRUE))

```

# `tapply()`
```{r, echo = FALSE}
options(width = 60)
```
\vspace{-4mm}
```{r, eval=FALSE}
with(combined.df, tapply(Baseline, Race.group, mytab))
```
\vspace{-9mm}
```{r, echo = FALSE}
with(combined.df, tapply(Baseline, Race.group, mytab))
```

# Data cleaning 
```{r} 
str(combined.df)
```

# Data cleaning 
```{r} 
str(combined.df[,11:13])
```

- a common problem is the dataset where some of the column names are not names of the variables, but *values* of variable. 
- the last three columns are Serum Cholesterol levels, mg/100ml, measured on: Day 1, Day 5 and Day 10, which we have recoded to `Baseline`, `PreTrt` and `PostTrt`, resepctively.  



# `tidyr` R package 

```{r, tidy = FALSE} 
library(tidyr)
```
Two main functions: 
    -  `gather()` - takes multiple columns and combines based at a *key* value, i.e. `Time`. 
    -  `spread()` - opposite of `gather()`

# `gather()` 

```{r}
combined.long.df <- gather(combined.df, "Baseline", "PreTrt", "PostTrt", key = Time, value= Cholesterol)

```

# `gather()` 

```{r, tidy = FALSE} 
combined.long.df <- gather(combined.df, key = Time,  value= Cholesterol, 
                           -ID, -Age, -Age.group, -Sex, -Weight, -Height,
                           -Smoke.group, -Race.group, -BMI,
                           -BMI.group)
```


# `gather()` 

```{r}
head(combined.long.df)
```

# `gather()`

```{r}
str(combined.long.df)
```


# `spread()` 

```{r}
combined.wide.df <- spread(combined.long.df, key = Time,
                           value = Cholesterol)

```

# `dplyr` R package 

```{r}

Patient.df <- read.csv("..\\..\\data\\Patient.csv", stringsAsFactors = FALSE)

step1 <- mutate(Patient.df, 
                Sex = as.factor(Sex), 
               Smoke.group = factor(ifelse(Smoke == 1, "Yes", "No")),
               Race.group =  factor(ifelse(Race == 1, "Caucasian",
              ifelse(Race == 2, "African", "Other"))), 
              BMI =  (Weight/Height^2)*703)

step2 <- select(step1, -Smoke, -Race)

step3 <- mutate(step2, 
               BMI.group =   factor(ifelse(BMI >= 30, "obese", 
         ifelse(BMI >= 25, "overweight",
          "normal")), 
         levels = c("normal", "overweight", "obese")))

step4 <- mutate(step3, 
               Age.group =   factor( ifelse(Age<=35, "Under 35", 
         ifelse(Age <= 60, "36 to 60", "Over 61")), 
         levels = c("Under 35", "36 to 60", "Over 61")))

Cholesterol.df <- read.csv("..\\..\\data\\Cholesterol.csv")

step5 <- left_join(step4, Cholesterol.df)

combine.wide.df <- rename(step5, ID = Patient.ID, Baseline = Day1, PreTrt = Day5, PostTrt = Day10)


```

# Piping `%>%` 

` x %>% f(y)` is equivalent to `f(x, y)`, where `x` is typically the data-frame.


# Piping operator `%>%` 

```{r, eval=FALSE, tidy=FALSE}
combine.wide.df <- Patient.df %>% mutate(
  Sex = as.factor(Sex),
  Smoke.group = factor(ifelse(Smoke == 1, "Yes", "No")),
  Race.group =  factor(ifelse(
    Race == 1, "Caucasian",
    ifelse(Race == 2, "African", "Other")
  )),
  BMI =  (Weight / Height ^ 2) * 703
) %>%
  select(-Smoke,-Race) %>%
  mutate(BMI.group = factor(
    ifelse(BMI >= 30, "obese",
           ifelse(BMI >= 25, "overweight",
                  "normal")),
    levels = c("normal", "overweight", "obese")
  )) %>%
 
  ```

# Piping operator `%>%` 

```{r, eval = FALSE, tidy=FALSE}
 mutate(Age.group =   factor(
    ifelse(Age <= 35, "Under 35",
           ifelse(Age <= 60, "36 to 60", "Over 61")),
    levels = c("Under 35", "36 to 60", "Over 61")
  )) %>%
  left_join(Cholesterol.df) %>% rename(
    ID = Patient.ID,
    Baseline = Day1,
    PreTrt = Day5,
    PostTrt = Day10
  )
 
 
combined.long.df <- combine.wide.df %>% 
  gather("Baseline", "PreTrt", "PostTrt",
         key = Time, value= Cholesterol)
```

# Grouping

```{r}
with(combined.df, tapply(Baseline, Race.group, mean, na.rm = TRUE))

combine.wide.df %>% group_by(Race.group) %>% 
  summarise(BaseMean = mean(Baseline, na.rm = TRUE))

```

# Grouping

```{r}
with(combined.df, tapply(Baseline, Race.group, mytab))

combine.wide.df %>% group_by(Race.group) %>% 
  summarise(Complete.obs = mytab(Baseline)[1],
            Missing.obs = mytab(Baseline)[2],
            Mean = mytab(Baseline)[1],
            Std.Error = mytab(Baseline)[3],
            Lower.CI = mytab(Baseline)[4],
            Upper.CI = mytab(Baseline)[5],
            )
```
# Grouping

```{r}

combine.wide.df %>% 
  group_by(Age.group, Race.group, BMI.group) %>% 
  count() 
```

# Grouping

```{r}

combine.wide.df %>% 
  group_by(Age.group, Race.group, BMI.group) %>% 
  count() %>% 
  arrange(desc(n))
```
# Grouping

```{r}
combine.wide.df %>% 
  group_by(Age.group, Race.group, BMI.group) %>% 
  count()  %>% 
  spread(Age.group, n) 
```

# Grouping

```{r}
combine.wide.df %>% 
  group_by(Age.group, Race.group, BMI.group) %>% 
  count()  %>% 
  spread(Age.group, n) %>% 
  filter(!is.na(BMI.group), !is.na(Race.group)) 

```

# Summary

- `apply()`
- `tapply()`
- `gather()` and `spread()` of `tidyr` R package
- `dplyr` R package
- piping operator `%>%`



