---
title: "Introduction to $\\mathtt{R}$"
header-includes: 
   \usepackage{graphicx}
   \usepackage{booktabs}
   \usepackage{caption}
    \subtitle{Session 3 -- Data manipulation}    
    \author[SCC]{Statistical Consulting Centre}
   \institute[\href{mailto:consulting@stat.auckland.ac.nz}
    {consulting@stat.auckland.ac.nz}]{\href{mailto:consulting@stat.auckland.ac.nz}
    {consulting@stat.auckland.ac.nz}\\
  The Department of Statistics\\
  The University of Auckland}
   \titlegraphic{\includegraphics[width=5cm]{..//..//S-DS-VC-RGB.png}}
date: "19 July, 2017"
output:
  beamer_presentation:
    theme: "Madrid"
    colortheme: "crane"
    fonttheme: "professionalfonts"
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(tidy = TRUE)
knitr::opts_chunk$set(message = FALSE)
options(width = 50)
Patient.df <- read.csv("..\\..\\data\\Patient.csv", stringsAsFactors = FALSE)
Patient.df$Smoke.group <- 
  with(Patient.df, ifelse(Smoke == 1, "Yes", "No"))
table(Patient.df$Smoke.group)
Patient.df$Race.group <- 
  with(Patient.df,
       ifelse(Race == 1, "Caucasian",
              ifelse(Race == 2, "African", "Other")))
table(Patient.df$Race.group)
Patient.df$BMI <- with(Patient.df, (Weight/Height^2)*703)
Patient.df$BMI.group <- 
  ifelse(Patient.df$BMI >= 30, "obese", 
         ifelse(Patient.df$BMI >= 25, "overweight",
          "normal"))

Patient.df <- Patient.df[, !names(Patient.df) %in% c("Smoke", "Race")]

```


# \texttt{factor}
  \begin{block}{What is a factor?}
    \textit{A variable which takes either qualitative values, ordinal values or a discrete set of quantitative values. The values of a factor are called its {\em levels}.}
  \end{block}
Examples of factors:  
\begin{itemize}
  \item \texttt{Gender} with 2 {\em qualitative} levels: \texttt{Male} and \texttt{Female}.
  \item \texttt{Education} with 6 {\em ordinal} levels: \texttt{None} $<$ \texttt{"Primary compl} $<$ \texttt{Incpl secondary} $<$ \texttt{Secondary compl} $<$ \texttt{Incpl university} $<$ \texttt{University degree}.
  \item \texttt{Income} has 9 {\em quantitative} levels when the mid-values of the income ranges are used: \texttt{5000}, \texttt{12500}, \texttt{17500}, \texttt{22500}, \texttt{27500}, \texttt{35000}, \texttt{45000}, \texttt{60000} and \texttt{85000}.
\end{itemize}

# \texttt{factor}

- \texttt{R} stores two *additional* pieces of information for each factor: (1) the unique set of levels and (2) an integer value, assigned by \texttt{R}, for each unique level. 
- The integer values are assigned to factor levels so that they have an order associated with them.
- By default, the unique levels are assigned the values 1, 2,..., according to ascending alphabetical order. This is not always appropriate! 

```{r}
typeof(Patient.df$Sex)
levels(Patient.df$Sex)
```

```{r}
Patient.df$BMI.group <- factor(Patient.df$BMI.group)
levels(Patient.df$BMI.group)

```

```{r}
Patient.df$BMI.group <- factor(Patient.df$BMI.group,
               levels = c("normal", "overweight", "obese"))
levels(Patient.df$BMI.group)

```

# Which other variables should also be factors?

```{r}
str(Patient.df)
```

# Which other variables should also be factors?

```{r}
Patient.df$Sex <- factor(Patient.df$Sex)
Patient.df$Race.group <- factor(Patient.df$Race.group)
Patient.df$Smoke.group <- factor(Patient.df$Smoke.group)
```


# Converting numbers to factors 

```{r}
test <- factor(c(0,1,2))
test
```
- then convert back to numbers? 

# And convert back to numbers

- need to convert it to character, using `as.character()` first, then convert back to numbers, using  `as.numeric()`. 
```{r}
test 

as.numeric(test)

as.numeric(as.character(test))
```

# Your turn (Binning ages into age groups)

- Sometimes we are interested in examining responses by age group.
- Now, assign each of the `r nrow(Patient.df)` patients to one of three age groups: "Under 35", "36 to 60" and "Over 61".
- Convert \texttt{Age.group} to a \texttt{factor} with levels in ascending order.


# Your turn 

Assign each of the `r nrow(Patient.df)` patients to one of three age groups: "Under 35", "36 to 60" and "Over 61".

```{r}
Patient.df$Age.group <- 
  ifelse(Patient.df$Age<=35, "Under 35", 
         ifelse(Patient.df$Age <= 60, "36 to 60", "Over 61"))
table(Patient.df$Age.group)

```

# Your turn 

Convert \texttt{Age.group} to a \texttt{factor} with levels in ascending order.

```{r}
Patient.df$Age.group <- factor(Patient.df$Age.group)
table(Patient.df$Age.group)

```

# Your turn 

Convert \texttt{Age.group} to a \texttt{factor} with levels in ascending order.

```{r}
Patient.df$Age.group <- factor(Patient.df$Age.group, 
                           levels = c("Under 35", "36 to 60", "Over 61"))
table(Patient.df$Age.group)

```


#Other way: `if`/`else` statement 

```{r, eval = FALSE}
  if(test) {
  #  yes
  } else{
  #  no
  }
```

\begin{itemize}
  \item \texttt{test}: a logical test.
  \item \texttt{yes}, what happens if the test is True.
  \item \texttt{no}, what happens if the test is False.
\end{itemize}

#Other way: `if`/`else` statement 

```{r, eval = FALSE}
  if(test) {
  #  yes for (test)
  } else if(test1) {
  #  no for (test) but yes for (test1) 
  } else{
  #  no for both (test) and (test1)
  }
```




# Relational data

- In general, it is rare in data analysis involves only a single table of data. 
- Examples: 
    + Patient information and blood test measurements
    + Experimental design and measurments from the high-thoughput biological instrument

# \texttt{Cholesterol.df}
Serum Cholesterol level, mg/100ml, measured on:

- `Day1` 
- `Day5`
- `Day10`

# Reading data into \texttt{R}
```{r, eval=FALSE, echo = TRUE}
setwd("your working directory")
Patient.df <- read.csv("CholesterolNA.csv")
head(Cholesterol.df)
```

```{r, eval=TRUE, echo = FALSE}
Cholesterol.df <- read.csv("..\\..\\data\\CholesterolNA.csv")
head(Cholesterol.df)
```

# \texttt{names()}, \texttt{dim()} and \texttt{str()}

```{r, eval=TRUE, echo = TRUE}
#Names of the variables
names(Cholesterol.df)
dim(Cholesterol.df)
str(Cholesterol.df)
```

# Combining data-frame by columns (\texttt{cbind()})
```{r}
combined.df <- cbind(Patient.df, Cholesterol.df[,-1])
```
- Thus, the function for combining the data-frame by rows is `rbind()`.
- Make sure the dimensions are correct for two combining data-frames. 
- Also, you need to make sure each row in `Patient.df` matches each row in `Cholesterol.df`.


# Reading data into \texttt{R} (Again)
```{r, eval=FALSE, echo = TRUE}
setwd("your working directory")
Patient.df <- read.csv("Cholesterol.csv")
head(Cholesterol.df)
```

```{r, eval=TRUE, echo = FALSE}
Cholesterol.df <- read.csv("..\\..\\data\\Cholesterol.csv")
head(Cholesterol.df)
```

# \texttt{names()}, \texttt{dim()} and \texttt{str()} (Again)

```{r, eval=TRUE, echo = TRUE}
#Names of the variables
names(Cholesterol.df)
dim(Cholesterol.df)
str(Cholesterol.df)
```


# Combining data-frame by columns (\texttt{cbind()})
```{r, error=TRUE}
combined.df <- cbind(Patient.df, Cholesterol.df)
```


# Better way 

- Combining the based on the `Patient.ID` in both `Patient.df` and `Cholesterol.df`.
- First thing is to make sure the `Patient.ID` are unique in both data-frames. 

# Better way 

- Combining the based on the `Patient.ID` in both `Patient.df` and `Cholesterol.df`.
- First thing is to make sure the `Patient.ID` are unique in both data-frames. 

```{r}
sum(table(Patient.df$Patient.ID) > 1)
sum(table(Cholesterol.df$Patient.ID) > 1)

```

# `dplyr` R package

- `dplyr` R package provides some useful functions that correspond to the most data manipulation tasks.

```{r}
library(dplyr)
```

- Mutating joins allow you to combine variables from multiple tables, there are four common types: 
    +  `left_join()`
    +  `right_join()`
    +  `full_join()`
    +  `inner_join()`


# `dplyr` R package
```{r, tidy=FALSE}
x <- data.frame(key = c(1,2,3), val.x = c("x1","x2","x3"))
y <- data.frame(key = c(1,2,4), val.y = c("y1","y2","y4"))
```

```{r, echo=FALSE, results='asis'}
t1 <- knitr::kable(x, format = "latex", booktabs = TRUE)
t2 <- knitr::kable(y, format = "latex", booktabs = TRUE)

cat(c("\\begin{table}[!htb]
    \\begin{minipage}{.5\\linewidth}
      \\caption*{Data-frame x}
      \\centering",
        t1,
    "\\end{minipage}%
    \\begin{minipage}{.5\\linewidth}
      \\centering
        \\caption*{Data-frame y}",
        t2,
    "\\end{minipage} 
\\end{table}"
)) 
```


# `left_join()` 
```{r, eval = FALSE, message=FALSE}
left_join(x, y, by = "key")
```


```{r, echo=FALSE, results='asis'}
t1 <- knitr::kable(x, format = "latex", booktabs = TRUE)
t2 <- knitr::kable(y, format = "latex", booktabs = TRUE)
t3 <- knitr::kable(left_join(x, y), format = "latex", booktabs = TRUE)

cat(c("\\begin{table}[!htb]
    \\begin{minipage}{.33\\linewidth}
      \\caption*{Data-frame x}
      \\centering",
        t1,
    "\\end{minipage}%
    \\begin{minipage}{.33\\linewidth}
      \\centering
        \\caption*{Data-frame y}",
        t2,
      "\\end{minipage}%
    \\begin{minipage}{.33\\linewidth}
      \\centering
        \\caption*{left join}",
        t3,
    "\\end{minipage} 
\\end{table}"
)) 
```


# `right_join()` 
```{r, eval = FALSE, message=FALSE}
right_join(x, y, by = "key")
```

```{r, echo=FALSE, results='asis'}
t1 <- knitr::kable(x, format = "latex", booktabs = TRUE)
t2 <- knitr::kable(y, format = "latex", booktabs = TRUE)
t3 <- knitr::kable(right_join(x, y), format = "latex", booktabs = TRUE)

cat(c("\\begin{table}[!htb]
    \\begin{minipage}{.33\\linewidth}
      \\caption*{Data-frame x}
      \\centering",
        t1,
    "\\end{minipage}%
    \\begin{minipage}{.33\\linewidth}
      \\centering
        \\caption*{Data-frame y}",
        t2,
      "\\end{minipage}%
    \\begin{minipage}{.33\\linewidth}
      \\centering
        \\caption*{right join}",
        t3,
    "\\end{minipage} 
\\end{table}"
))  
```

# `full_join()` 
```{r, eval = FALSE, message=FALSE}
full_join(x, y, by = "key")
```


```{r, echo=FALSE, results='asis'}
t1 <- knitr::kable(x, format = "latex", booktabs = TRUE)
t2 <- knitr::kable(y, format = "latex", booktabs = TRUE)
t3 <- knitr::kable(full_join(x, y), format = "latex", booktabs = TRUE)

cat(c("\\begin{table}[!htb]
    \\begin{minipage}{.33\\linewidth}
      \\caption*{Data-frame x}
      \\centering",
        t1,
    "\\end{minipage}%
    \\begin{minipage}{.33\\linewidth}
      \\centering
        \\caption*{Data-frame y}",
        t2,
      "\\end{minipage}%
    \\begin{minipage}{.33\\linewidth}
      \\centering
        \\caption*{full join}",
        t3,
    "\\end{minipage} 
\\end{table}"
))  
```

# `inner_join()` 
```{r, eval = FALSE, message=FALSE}
inner_join(x, y, by = "key")
```

```{r, echo=FALSE, results='asis'}
t1 <- knitr::kable(x, format = "latex", booktabs = TRUE)
t2 <- knitr::kable(y, format = "latex", booktabs = TRUE)
t3 <- knitr::kable(inner_join(x, y), format = "latex", booktabs = TRUE)

cat(c("\\begin{table}[!htb]
    \\begin{minipage}{.33\\linewidth}
      \\caption*{Data-frame x}
      \\centering",
        t1,
    "\\end{minipage}%
    \\begin{minipage}{.33\\linewidth}
      \\centering
        \\caption*{Data-frame y}",
        t2,
      "\\end{minipage}%
    \\begin{minipage}{.33\\linewidth}
      \\centering
        \\caption*{inner join}",
        t3,
    "\\end{minipage} 
\\end{table}"
)) 
```

# `dplyr` R package

These four types of mutating join are differ in their behaviour when a match is not found. 

-  `left_join()`
-  `right_join()`
-  `full_join()`
-  `inner_join()`

Which one should we use for this `Patient.df` and `Cholesterol.df` data-frames?

# Combining two tables

```{r}
combined.df <- left_join(Patient.df, Cholesterol.df)
```

- Variable names:
```{r}
names(combined.df)
```

# Combining two tables

```{r}
str(combined.df)
```

# Combining two tables
You can change the variable names.
```{r, tidy=FALSE}
names(combined.df)[c(1,11,12,13)] <- 
  c("ID", "Baseline", "PreTrt", "PostTrt")
names(combined.df)
```

# Categorical Variables


```{r}
table(combined.df$Age.group)
table(combined.df$Sex)
```
.
.
.



# `for` loop to get column summary statistics
```{r, eval = FALSE}
for (i in c("Age.group", "Sex", "Smoke.group", "Race.group", "BMI.group")){
  print(i)
  print(table(combined.df[,i]))
}
```


# Continous Variables

```{r}
mean(combined.df$Age)

```



# `for` loop to get column summary statistics
```{r}
for (i in c("Age", "Height", "Weight", "BMI", "Baseline", "PreTrt", "PostTrt")){
  print(i)
  print(mean(combined.df[,i]))
}
```

# `for` loop to get column summary statistics
```{r}
for (i in c("Age", "Height", "Weight", "BMI", "Baseline", "PreTrt", "PostTrt")){
  print(i)
  print(mean(combined.df[,i], na.rm = TRUE))
}
```

#Easier way 

```{r}
summary(combined.df[,-1])

```


# Summary

- `factor`
- Joining the data-sets
- `for` loop











