---
title: "Introduction to $\\mathtt{R}$"
subtitle: "Session 5 -- Advanced Graphics"
author: "Statistical Consulting Centre"
date: "20 July, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = FALSE)
# knitr::opts_chunk$set(eval = FALSE)

lake.df <- read.csv("..\\..\\data\\Lake.csv", 
                      stringsAsFactors = FALSE)

lake.df$pHtype <- with(lake.df, ifelse(pH > 7, "alkalinity", ifelse(pH < 7, "acidity", "natural")))
lake.df$pHtype <- factor(lake.df$pHtype, levels = c( "acidity", "natural", "alkalinity"))
lake.df$Calcium <- factor(lake.df$Calcium, levels = c("Low", "Medium", "High"))

mercury.df <- read.csv("..\\..\\data\\Mercury.csv", 
                      stringsAsFactors = FALSE)
library(dplyr)
joined.df <- left_join(lake.df, mercury.df, by = "ID")

library(tidyr)
joined.long.df <- gather(joined.df, "Day1", "Day2", "Day3", key = Time, value= Mercury)

library(ggplot2)

```


# 1. T-test

1. Do a two-sample $t$-test (two-sided) to see whether \texttt{Mercury} for acidic lakes is different from other \texttt{pHtype} levels, assuming 

  (a) population variances are equal, and

```{r}
joined.long.df$pHtypeAcidic <-  with(joined.long.df, ifelse(pH < 7, "acidity", "Others"))

with(joined.long.df, t.test(Mercury~pHtypeAcidic, var.equal = TRUE))
```

 (b) population variances are not equal

```{r}

with(joined.long.df, t.test(Mercury~pHtypeAcidic, var.equal = FALSE))
```

Hint: Create a new variable \texttt{pHtypeAcidic} for just acidic group.

2. How do the results from using the following code compare with the results you got in question 1?

````
acid.mer <- with(joined.long.df, mercury[pHtype == "acidity"])
not.acid.mer <- with(joined.long.df, mercury[pHtype != "acidity"])

t.test(acid.mer, not.acid.mer)

````

```{r}

acid.mer <- with(joined.long.df, Mercury[pHtype == "acidity"])
not.acid.mer <- with(joined.long.df, Mercury[pHtype != "acidity"])

t.test(acid.mer, not.acid.mer)

```


# 2. One-way ANOVA

1. Perform an ANOVA to test the null hypothesis that the mean mercury concentrations are all equal for all three Calcium levels.

```{r}

myaov <- with(joined.long.df, aov(Mercury ~ Calcium))

```

2. Use `summary()` to check the overall signifcance of `Calcium`.

```{r}

summary(myaov)

```


3. Use `model.tables()` to calculate the mean mercury concentrations for each Calcium level.


```{r}
model.tables(myaov, type = "means")
```


4. Carry out pairwise comparisons of mean mercury concentrations between pairs of Calcium levels, adjusting the p-values using Tukey's Honestly Signifcance Difference.

```{r}
TukeyHSD(myaov)

```


# 3. Two-way ANOVA

1. Perform an Two-way ANOVA to test the null hypothesis that the mean mercury concentrations are all equal for all three Calcium levels across all three days.

```{r}

myaov <- with(joined.long.df, aov(Mercury ~ Calcium*Time))

```

2. Use `summary()` to check the overall signifcance.

```{r}

summary(myaov)

```


3. Use `model.tables()` to calculate the mean mercury concentrations for each Calcium level.


```{r}
model.tables(myaov, type = "means")
```


4. Carry out pairwise comparisons of mean mercury concentrations, adjusting the p-values using Tukey's Honestly Signifcance Difference.

```{r}
TukeyHSD(myaov)

```




# 4. Test of independence

1. Perform a Pearson's Chi-squared test to check if the Calcium depend on pHtype level? 

```{r}
ca.ph.tab <- with(joined.df, table(Calcium, pHtype))
ca.ph.tab

chisq.test(ca.ph.tab)

```

2. Why there is a warning from the 4.1? What is an altertive test which we can perform?


```{r} 
fisher.test(ca.ph.tab)

```


# 5. Linear regression

1. Produce a scatterplot between pH (x-axis) and mercury (y-axis).

```{r}
with(joined.long.df, plot(pH, Mercury))

```


2. Describe the relationship between pH and mercury showing in the scatterplot.


3. Fit a linear model between pH and mercury, in which mercury is the dependent (response) variable and pH is the independent (explanatory) variable. Name the linear model as `mylm`.


```{r}
mylm <- with(joined.long.df, lm(Mercury ~ pH))
```

4. Use `summary()` to obtain the estimates and test the statistical signifcance of the intercept and slope.

```{r}
summary(mylm)
```

5 Write down the equation of the fitted line.


6. Use the following code to check the residuals, for homogeneity of variance.

````
plot(predict(mylm), residuals(mylm))
abline(h = 0, lwd = 2, col = 2)
````

```{r}
plot(predict(mylm), residuals(mylm))
abline(h = 0, lwd = 2, col = 2)
```








