---
title: "NZSSN Courses: Introduction to $\\mathtt{R}$"
header-includes: 
   \usepackage{graphicx}
    \subtitle{Session 7 -- Simple analysis}    
    \author[SCC]{Statistical Consulting Centre}
   \institute[\href{mailto:consulting@stat.auckland.ac.nz}
    {consulting@stat.auckland.ac.nz}]{\href{mailto:consulting@stat.auckland.ac.nz}
    {consulting@stat.auckland.ac.nz}\\
  The Department of Statistics\\
  The University of Auckland}
   \titlegraphic{\includegraphics[width=5cm]{..//..//S-DS-VC-RGB.png}}
date: "20 July, 2017"
output:
  beamer_presentation:
    theme: "Madrid"
    colortheme: "crane"
    fonttheme: "professionalfonts"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy = FALSE)
knitr::opts_chunk$set(message = FALSE)

options(width = 60)
library(dplyr)
library(tidyr)

Patient.df <- read.csv("..\\..\\data\\Patient.csv", stringsAsFactors = FALSE)
Patient.df$Smoke.group <- 
  with(Patient.df, ifelse(Smoke == 1, "Yes", "No"))
table(Patient.df$Smoke.group)
Patient.df$Race.group <- 
  with(Patient.df,
       ifelse(Race == 1, "Caucasian",
              ifelse(Race == 2, "African", "Other")))
table(Patient.df$Race.group)
Patient.df$BMI <- with(Patient.df, (Weight/Height^2)*703)
Patient.df$BMI.group <- 
  ifelse(Patient.df$BMI >= 30, "obese", 
         ifelse(Patient.df$BMI >= 25, "overweight",
          "normal"))

Patient.df <- Patient.df[, !names(Patient.df) %in% c("Smoke", "Race")]

Patient.df$BMI.group <- factor(Patient.df$BMI.group,
               levels = c("normal", "overweight", "obese"))

Patient.df$Sex <- factor(Patient.df$Sex)
Patient.df$Race.group <- factor(Patient.df$Race.group)
Patient.df$Smoke.group <- factor(Patient.df$Smoke.group)

Patient.df$Age.group <- 
  ifelse(Patient.df$Age<=35, "Under 35", 
         ifelse(Patient.df$Age <= 60, "36 to 60", "Over 61"))
Patient.df$Age.group <- factor(Patient.df$Age.group, 
                           levels = c("Under 35", "36 to 60", "Over 61"))
Cholesterol.df <- read.csv("..\\..\\data\\CholesterolNA.csv")

library(dplyr)

combined.df <- left_join(Patient.df, Cholesterol.df)

names(combined.df)[c(1,11,12,13)] <- 
  c("ID", "Baseline", "PreTrt", "PostTrt")

combined.long.df <- gather(combined.df, key = Time,  value= Cholesterol, 
                           -ID, -Age, -Age.group, -Sex, -Weight, -Height,
                           -Smoke.group, -Race.group, -BMI,
                           -BMI.group)
```


# Regression commands

Two of the most commonly used \texttt{R} commands for modeling:

- \texttt{lm()}: fits \textbf{L}inear \textbf{M}odels
- \texttt{glm()}: fits \textbf{G}eneralised \textbf{L}inear \textbf{M}odels.\\\vspace{0.5cm}

  Note SAS users: \texttt{PROC GLM} is \textbf{not} the same as \texttt{R}'s \texttt{glm()}.


There's a lot in these two commands; entire stage 3 statistical courses on linear and generalised linear models.

# Student's $t$-test
  \begin{center}
  \texttt{t.test(y $\sim$ x)}
  \end{center}

- \texttt{y}: values; e.g., \texttt{Cholesterol, BMI, Age}, etc.
- \texttt{x}: group; e.g., \texttt{Sex, Smoke.group}.

Suppose we want to test whether males and females (\texttt{x = Sex}) have different Cholesterol levels.

Categorical variables should be converted to type \texttt{factor} before analysis, i.e.

```{r, eval=FALSE}
combined.long.df$Sex <- factor(combined.long.df$Sex)
with(combined.long.df, t.test(Cholesterol ~ Sex))
```

# Student's $t$-test

```{r, echo=FALSE}
combined.long.df$Sex <- factor(combined.long.df$Sex)
with(combined.long.df, t.test(Cholesterol ~ Sex))
```

- p-value < 2.2e-16.
- We have extremely strong evidence that the cholesterol level for male is different from female.


# Multiple comparisons
Let's compare the total score between three age groups, i.e.
\begin{enumerate}
\item Do a $t$-test between "Under 35" and "36 to 60".
\item Do a $t$-test between "Under 35" and "Over 61".
\item Do a $t$-test between "36 to 60" and "Over 61".
\end{enumerate}

\begin{center}
\huge{Really?}
\end{center}


# Error rate
When we do a $t$-test comparing mean total score between females and males, the null hypothesis is that the mean total score for females is the same as that for males. The $t$-test is performed (with the hope) to reject this null hypothesis.

In order to come up with a p-value, we \emph{assume} that $\alpha$ (typically 5\%) of the time, we will reject the null hypothesis when it's actually true, i.e., we assume 5\% of the time we will make a mistake.

\begin{itemize}
\item When we do two simultaneous $t$-tests, about 10\% of the time we will make a mistake.
\item When we do three simultaneous $t$-tests, about 15\% of the time we will make a mistake.
\item The chance of being shot in Russian Roulette is 16.67\%. Would
  you risk it then?
\end{itemize}


# \textit{\textbf{An}}alysis \textit{\textbf{o}}f \textit{\textbf{Va}}riance (ANOVA)
Generalises $t$-test to more than two groups

Null hypothesis: all group means are equal.

**Example.** Mean Cholesterol level is the same for all three age.groups.

```{r}
tryaov <- with(combined.long.df, aov(Cholesterol~Age.group))
```

\begin{itemize}
  \item \texttt{aov()}: \textbf{A}nalysis \textbf{o}f \textbf{V}ariance.
  \item Response variable (i.e. \texttt{total.lik}) is separated by \texttt{\~} from explanatory variable(s) (i.e. \texttt{age.group}).
  \item All explanatory variables should be categorical (otherwise it's not ANOVA).
\end{itemize} 


#\texttt{aov()}

```{r}
summary(tryaov)
```

We have extremely strong evidence that at least one age group's mean Cholesterol level is different to that of the other age groups.\\\vspace{0.3cm}

Which one(s) is(are) different????

# Which one(s)?
```{r}
model.tables(tryaov, "means")
```

```{r echo = FALSE}
m <- round(model.tables(tryaov, "means")$tables$'Grand mean', 1)
gm <- round(model.tables(tryaov, "means")$tables$Age.group, 1)
```

The mean Cholesterol level...
\begin{itemize}
\item over all participants is `r m`.
\end{itemize}

# Which one(s)?
```{r}
model.tables(tryaov, "means")
```

```{r echo = FALSE}
m <- round(model.tables(tryaov, "means")$tables$'Grand mean', 1)
gm <- round(model.tables(tryaov, "means")$tables$Age.group, 1)
```

The mean Cholesterol level...
\begin{itemize}
\item for "Under 35" group is lower than both that of the "36 to 60" and the "Over 61" groups.
\item for "36 to 60" group is lower than the "Over 61" group.
\end{itemize}


# Which one(s)?
```{r}
model.tables(tryaov, "means")
```

```{r echo = FALSE}
m <- round(model.tables(tryaov, "means")$tables$'Grand mean', 1)
gm <- round(model.tables(tryaov, "means")$tables$Age.group, 1)
```

Are any pairs of these means statistically different from one another?

# Post-hoc multiple comparisons
\vspace{-4mm}
```{r}
TukeyHSD(tryaov)
```
\vspace{-6mm}
\begin{itemize}
\item \texttt{diff}: estimated difference between two group means.
\item \texttt{lwr, upr}: lower and upper limit of the 95\% confidence interval of the estimated difference.
\item \texttt{p adj}: \texttt{p}-values adjusted for multiple comparisons.
\end{itemize}

# Post-hoc multiple comparisons
```{r}
comp <- TukeyHSD(tryaov)
comp$Age.group
```
\begin{itemize}
  \item Mean Cholesterol level for "36 to 60" is `r abs(round(comp$Age.group[1, 1], 1))` mg/100ml \emph{higher} than "Under 35" (\texttt{p adj} $<$ 0.0001).
  \item Mean Cholesterol level for "Over 61" is `r abs(round(comp$Age.group[2, 1], 1))` mg/100ml \emph{higher} than "Under 35" (\texttt{p adj} $<$ 0.0001).
  \item Mean Cholesterol level for "Over 61" is `r abs(round(comp$Age.group[3, 1], 1))` mg/100ml \emph{higher} than "36 to 60" (\texttt{p adj} $<$ 0.0001).
\end{itemize}

# From Session 6: Mean Cholesterol level vs Age group
\begin{center}
  Mean $\pm$ 1.96 $\times$ SEM
\end{center}
```{r, echo = FALSE} 
pp <- data.frame(comp$Age.group)
pp$Age.group <- factor(rownames(comp$Age.group))

library(ggplot2)
ggplot(pp, aes(x = Age.group, y = diff)) + geom_point() + geom_errorbar(aes(ymax = upr, ymin = lwr))
```



# Two-way ANOVA

- \texttt{tryaov} was fitted using one categorical explanatory variable (\texttt{Age.group}). We therefore refer to its ANOVA table as \emph{one-way}.
- If we fit a linear model using two categorical explanatory variables, we have a \emph{two-way} ANOVA.
- Recall: All categorical variables should be converted into factors.
  
```{r, tidy = FALSE}
combined.long.df$Sex <- factor(combined.long.df$Sex)
try2way <- with(combined.long.df, 
                aov(Cholesterol~Sex*Age.group))
```
  
 - `Sex*Age.group` is equivalent to `Sex + Age.group + Sex:Age.group`.


# Two-way ANOVA
```{r}
summary(try2way)
```
There is two-way interaction between \texttt{Sex} and \texttt{Age.group} ($p$-value = 0.19), i.e., the magnitude of the difference in mean Cholesterol levels between males and females is not constant across all age groups, and vice versa.


# Estimated means
```{r}
model.tables(try2way, "means")
```

# Estimated means
```{r}
model.tables(try2way, "means")$table$'Sex:Age.group'
```


# Post-hoc pairwise comparisons
```{r}
TukeyHSD(try2way)
```

# Post-hoc pairwise comparisons
```{r}
TukeyHSD(try2way)$`Sex:Age.group`
```


# Test of independence
```{r}
smoke.age.tab <- with(combined.df, table(Smoke.group, Age.group))
smoke.age.tab
```

Do smoking habit depend on age group? 

Statistically speaking, is \texttt{Smoke.group} and \texttt{Age.group} independent of one another?


# Pearson's Chi-squared test
```{r}
chisq.test(smoke.age.tab)
```
\begin{itemize}
  \item There is extremely strong evidence (\texttt{p}-value $<$ 0.0001) that \texttt{Smoke.group} and \texttt{Age.group} are not independent of one another.
  
  \item Smoking habit depend on the age group to which patient belong.
\end{itemize}


# Assumptions

- Pearson's Chi-squared tests have certain assumptions. Beyond the scope of this course.
- \texttt{chisq.test()} will give you a warning if these assumptions are not met.

```{r echo = FALSE, error=TRUE}
mytest = 1:4
chisq.test(mytest)
```

- These assumptions are more likely to be wrong if the sample size is small.
- If this happens, the alternative is to use Fisher's exact test.


# Fisher's exact test
Assume \texttt{Q5.age.tab} does not meet the underlying assumptions of Pearson's Chi-squared test.

```{r}
fisher.test(smoke.age.tab, simulate.p.value = TRUE)
```

# Linear regression

`lm(y~x)` is used for linear regression.
\begin{itemize}
  \item \texttt{y}, the response variable.
  \item \texttt{x}, the explanatory variable.   
  \item There can be more than one explanatory variable, called {\em multiple} linear regression. 
  \item Both response variable and explanatory variable(s) should be numeric, it is {\em generalised} linear regression.
\end{itemize}


# Simple linear regression
When there is only one predictor variable (e.g. Age) in our linear regression, we refer to this as  *simple* linear regression.

```{r, eval = FALSE, tidy = FALSE}
with(combined.long.df, plot(Age, Cholesterol))
```

# Simple linear regression
```{r, echo=FALSE}
with(combined.long.df, plot(Age, Cholesterol))
```

# Normality check

```{r}
qqnorm(combined.long.df$Cholesterol)
qqline(combined.long.df$Cholesterol, col = 2, lwd = 2)
```

# Normality check

```{r}
qqnorm(log(combined.long.df$Cholesterol))
qqline(log(combined.long.df$Cholesterol), col = 2, lwd = 2)
```


# Simple linear regression

Let's carry out the linear regession of Age on Cholesterol level, i.e.

```{r, eval = FALSE, tidy = FALSE}
trylm <- with(combined.long.df, lm(log(Cholesterol)~Age))
summary(trylm)
```

# Simple linear regression

```{r, echo = FALSE}
trylm <- with(combined.long.df, lm(log(Cholesterol)~Age))
summary(trylm)
```

# Simple linear regression

```{r, echo = FALSE}
round(summary(trylm)$coef, 4)
```


- The estimated intercept is 5.1339. There is a very strong evidecne that this is not zero ($p$-value < 0.0001). 
- The estimated slope is 0.0035. There is a vereys strong evidence that this is not zero ($p$-value < 0.0001).
- The fitted line is:

$$ log(Cholesterol) = 5.1339 + 0.0035 \times Age $$
$$ Cholesterol = e^{5.1339 + 0.0035 \times Age} $$

- For every one year increase in age, the Cholesterol level increase by 1.003506 times.  

#Check 

```{r}
plot(residuals(trylm))

```

# Conclusion

- The linear relationship between age and Cholesterol level is statistically significant.
- Cholesterol level is positive related to age.

# Add the fitted line

```{r}
with(combined.long.df, plot(Age, log(Cholesterol)))
abline(trylm, col = 2)
```

# Quadratic term

```{r, tidy=FALSE}
tryquad <- with(combined.long.df,
        lm(log(Cholesterol) ~ Age + I(Age^2)))
round(summary(tryquad)$coef, 4)
```

`I(Age^2)` tells R to treat `^` as arithmetical operator, rather than formula operator.

Our fitted curve is:

$$ log(Cholesterol) = 4.8649 + 0.0035 \times Age + -0.0001Age^2 $$

# Summary

\begin{itemize}
  \item Student's $t$-test
  \item One-way ANOVA
  \item Two-way ANOVA
  \item Pearson’s Chi-squared test
  \item Fisher’s exact test
  \item linear regression
\end{itemize}










































































