---
title: "NZSSN Courses: Introduction to $\\mathtt{R}$"
header-includes: 
   \usepackage{graphicx}
    \subtitle{Session 6 -- Advanced Graphics}    
    \author[SCC]{Statistical Consulting Centre}
   \institute[\href{mailto:consulting@stat.auckland.ac.nz}
    {consulting@stat.auckland.ac.nz}]{\href{mailto:consulting@stat.auckland.ac.nz}
    {consulting@stat.auckland.ac.nz}\\
  The Department of Statistics\\
  The University of Auckland}
   \titlegraphic{\includegraphics[width=5cm]{..//..//S-DS-VC-RGB.png}}
date: "20 July, 2017"
output:
  beamer_presentation:
    theme: "Madrid"
    colortheme: "crane"
    fonttheme: "professionalfonts"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy = FALSE)
knitr::opts_chunk$set(message = FALSE)

options(width = 50)
library(dplyr)
library(tidyr)

Patient.df <- read.csv("..\\..\\data\\Patient.csv", stringsAsFactors = FALSE)
Patient.df$Smoke.group <- 
  with(Patient.df, ifelse(Smoke == 1, "Yes", "No"))
table(Patient.df$Smoke.group)
Patient.df$Race.group <- 
  with(Patient.df,
       ifelse(Race == 1, "Caucasian",
              ifelse(Race == 2, "African", "Other")))
table(Patient.df$Race.group)
Patient.df$BMI <- with(Patient.df, (Weight/Height^2)*703)
Patient.df$BMI.group <- 
  ifelse(Patient.df$BMI >= 30, "obese", 
         ifelse(Patient.df$BMI >= 25, "overweight",
          "normal"))

Patient.df <- Patient.df[, !names(Patient.df) %in% c("Smoke", "Race")]

Patient.df$BMI.group <- factor(Patient.df$BMI.group,
               levels = c("normal", "overweight", "obese"))

Patient.df$Sex <- factor(Patient.df$Sex)
Patient.df$Race.group <- factor(Patient.df$Race.group)
Patient.df$Smoke.group <- factor(Patient.df$Smoke.group)

Patient.df$Age.group <- 
  ifelse(Patient.df$Age<=35, "Under 35", 
         ifelse(Patient.df$Age <= 60, "36 to 60", "Over 61"))
Patient.df$Age.group <- factor(Patient.df$Age.group, 
                           levels = c("Under 35", "36 to 60", "Over 61"))
Cholesterol.df <- read.csv("..\\..\\data\\CholesterolNA.csv")

library(dplyr)

combined.df <- left_join(Patient.df, Cholesterol.df)

names(combined.df)[c(1,11,12,13)] <- 
  c("ID", "Baseline", "PreTrt", "PostTrt")

combined.long.df <- gather(combined.df, key = Time,  value= Cholesterol, 
                           -ID, -Age, -Age.group, -Sex, -Weight, -Height,
                           -Smoke.group, -Race.group, -BMI,
                           -BMI.group)
```
##`ggplot2` package
- Documentation: 
[http://docs.ggplot2.org/current/](http://docs.ggplot2.org/current/)

- recommended reading "The Layered Grammar of Graphics":
[http://vita.had.co.nz/papers/layered-grammar.pdf](http://vita.had.co.nz/papers/layered-grammar.pdf)

- Load `ggplot2` package
```{r}
library(ggplot2)
```

## Create a new ggplot

- Initialising a ggplot object.
```{r, fig.height=2, fig.width=3, warning=FALSE, fig.align='center'}
ggplot(data = combined.long.df,
       mapping = aes(x = Age, y = Cholesterol))
```

## Create a new ggplot

- Initialising a ggplot object.
```{r, eval=FALSE}
ggplot(data = combined.long.df, 
       mapping = aes(x = Age, y = Cholesterol))
```
There are three common ways to invoke ggplot:

- `ggplot(combined.long.df, aes(x, y, <other aesthetics>))`
- `ggplot(combined.long.df)`
- `ggplot()`

## Create a new ggplot
- assign this ggplot object to a variable
```{r, fig.height=2, fig.width=3, warning=FALSE, fig.align='center'}
g <- ggplot(data = combined.long.df, 
       mapping = aes(x = Age, y =  Cholesterol))
```

## Create a Scatterplot
```{r, fig.height=2, fig.width=4, warning=FALSE, fig.align='center'}
g + geom_point()
```
- `geom`, short for geometric object, describes the type of plot you will produce.

## Create a Scatterplot
- Note that here are three common ways to invoke ggplot: 
```{r, eval=FALSE, tidy = FALSE}
ggplot(data = combined.long.df, 
       mapping = aes(x = Age, y =  Cholesterol)) + geom_point()
ggplot(data = combined.long.df) + 
  geom_point( mapping = aes(x = Age, y =  Cholesterol))
ggplot() + 
  geom_point(data = combined.long.df, 
             mapping = aes(x = Age, y =  Cholesterol))
```
- always check the documentation, `?geom_point`, for which aesthetics can be used. 

## First method is recommended
```{r, fig.height=2, fig.width=3, warning=FALSE, fig.align='center'}
g <- ggplot(data = combined.long.df, 
       mapping = aes(x = Age, y =  Cholesterol))
g + geom_point()
```

## Control colour and shape
```{r, fig.height=2, fig.width=4, warning=FALSE, fig.align='center'}
g + geom_point(aes(colour = Sex, shape = Smoke.group))
```
- always check the documentation, `?geom_point`, for which aesthetics can be used.
- note the missing values in the legend labelling

## Modify axis, legend, and plot labels
```{r, fig.height=2.25, fig.width=4, warning=FALSE, fig.align='center'}
(g <- g + geom_point(aes(colour = Sex, shape = Smoke.group)) +
   labs(title = "Cholesterol level versus Age",
   x = "Age", y = "Cholesterol level (mg/100ml)"))
```

## Theme controls non-data components of the plot
```{r, fig.height=2, fig.width=4, warning=FALSE, fig.align='center'}
g + theme(plot.title = element_text(size=8, hjust = 0.5),
        axis.title = element_text(size=8),
        legend.title = element_text(size=8), 
        legend.text = element_text(size=8) )
```

## Create a Scatterplot
```{r, fig.height=2, fig.width=4, warning=FALSE, fig.align='center'}
g <- ggplot(
  data = na.omit(combined.long.df[,
              c("Age", "Cholesterol", "Sex", "Smoke.group")]), 
  mapping = aes(x = Age, y =  Cholesterol, 
                color = Sex, shape = Smoke.group)) + 
  geom_point() +
  labs(title ="Total score versus Age",  
        x = "Age", y = "Total score")
```

- `ggplot` object can be further modified. 
 

## Create a Scatterplot
```{r, fig.height=2, fig.width=4, warning=FALSE, fig.align='center'}
myTheme <- theme_bw() +
  theme(plot.title = element_text(size=8, hjust = 0.5),
        axis.title=element_text(size=8),
        legend.title = element_text(size=8),
        legend.text = element_text(size=8) )
```

- `myTheme` can be reused for different types of plot.


## Scatterplot
```{r, fig.height=2.5, fig.width=5, warning=FALSE, fig.align='center'}
g +  myTheme
```

## Scatterplot with a smoother
```{r, fig.height=2.5, fig.width=5, warning=FALSE, fig.align='center', message=FALSE}
g +  geom_smooth() + myTheme
```

## Bar chart
```{r, fig.height=2.5, fig.width=4, warning=FALSE, fig.align='center'}
ggplot(combined.long.df, aes(x = Age.group)) + geom_bar() + myTheme
```

##  Smoking group by Age group
```{r, fig.height=2, fig.width=5, warning=FALSE, fig.align='center'}
ggplot(na.omit(combined.long.df[,c("Age.group", "Smoke.group")]),
       aes(x = Age.group, fill = Smoke.group)) +
      geom_bar() + myTheme
```

## Bar chart on proportions

```{r, fig.height=2, fig.width=5, warning=FALSE, fig.align='center'}
g <- ggplot(na.omit(combined.long.df[,c("Age.group", "Smoke.group")]),
  aes(x = Age.group, y = (..count..)/sum(..count..),
      fill = Smoke.group))
g + geom_bar() + myTheme
```

## Bar chart on proportions
```{r, fig.height=2, fig.width=5, warning=FALSE, fig.align='center'}
g + geom_bar(position = "dodge") +
  labs( x = "Age group in years", y = "Percentage")+
  myTheme
```

## Boxplot using geom_boxplot()
```{r , fig.height=2, fig.width=3, warning=FALSE, fig.align='center'}
ggplot(na.omit(combined.long.df[,c("Age.group", "Cholesterol")]),
       aes(x = Age.group, y = Cholesterol)) +
  geom_boxplot()  +
  labs(title = "Cholesterol level versus Age",
      x = "Age Group", y = "Cholesterol level (mg/100ml)") +  myTheme
```

## Boxplot with panels using facet_wrap()
```{r, fig.height=3, fig.width=4, warning=FALSE, fig.align='center'}
g <- ggplot(na.omit(combined.long.df[,c("Age.group",
                               "Cholesterol", "Race.group")]),
       aes(x = Age.group, y = Cholesterol)) +
  geom_boxplot()  + labs(title = "Total Score versus Age",
            x = "Age Group", y = "Total score")
```

## Boxplot with panels using facet_wrap()
```{r}
g + facet_wrap(~Race.group) + myTheme
```


## Histogram

```{r, fig.height=2, fig.width=4, warning=FALSE, fig.align='center', message = FALSE}
ggplot(combined.long.df, aes(x = Age)) +
  geom_histogram() + myTheme

```

## Histogram with wider binwidth

```{r, fig.height=2, fig.width=4, warning=FALSE, fig.align='center'}
ggplot(combined.long.df, aes(x = Age)) +
  geom_histogram(binwidth = 10)+
  labs(x = "Age") + myTheme
```

## Histogram with narrower binwidth

```{r, fig.height=2, fig.width=4, warning=FALSE, fig.align='center'}
ggplot(combined.long.df, aes(x = Age)) +
  geom_histogram(binwidth = 1)+
  labs(x = "Age")  + myTheme
```

## Plot means in context

```{r}
with(combined.long.df, tapply(Cholesterol, Age.group,
                     mean, na.rm = TRUE))
```
- Means are all but meaningless unless they are presented in context.
- Always present with standard deviations (SDs) or standard error of means (SEs) or confidence intervals.
- Plot means with 95\% confidence intervals ($\pm$ 1.96 $\times$ SE).
    + $\pm$ 1 $\times$ SE yields (approx.) a `r round((1 - pnorm(-1)*2)*100, 0)`\% confidence interval. Equivalent to using a 16\% level of significance!!!!
    + $\pm$ 1 $\times$ SD tells us \textbf{ABSOLUTELY NOTHING} about whether two means are statistically different from one another.


## Calculating 95\% CIs
-  95\% CI $=$ Mean $\pm$ 1.96 $\times$ SE
- Standard Errors $= \frac{\text{Standard Deviation}}{\sqrt{\text{Sample Size}}}$
```{r}
my.m <- with(combined.long.df, tapply(Cholesterol, Age.group, mean,
                             na.rm = TRUE))
my.m
my.sd <- with(combined.long.df, tapply(Cholesterol, Age.group, sd,
                              na.rm = TRUE))
my.sd
```


## Calculating 95\% CIs
```{r}
my.n <- with(combined.long.df, tapply(Cholesterol, Age.group,
            function(x)length(which(!is.na(x)))))
my.n
my.stder <- my.sd/sqrt(my.n)
ci.upper <- my.m + 1.96*my.stder
ci.lower <- my.m - 1.96*my.stder
```

## Calculating 95\% CIs
```{r}
my.stder <- my.sd/sqrt(my.n)
ci.upper <- my.m + 1.96*my.stder
ci.lower <- my.m - 1.96*my.stder

cbind(my.m, ci.lower, ci.upper)

```

## Calculating 95\% CIs
```{r}
y.df = data.frame(age.group =  factor(names(my.m)),
                  my.m, ci.upper, ci.lower)

y.df
```

## Errorbars

```{r, eval=FALSE}
ggplot(y.df, aes(x = age.group, y = my.m))+geom_point() +
     geom_errorbar(aes(ymax = ci.upper, ymin = ci.lower),
                   width = 0.1)+
     xlab("Age Group")+
     ylab("Mean total feminist score") +
  myTheme
```




## Errorbars

```{r, fig.height=3, fig.width=4, warning=FALSE, fig.align='center' , echo=FALSE}
y.df = data.frame(age.group =  factor(names(my.m)),
                  my.m, ci.upper, ci.lower)

ggplot(y.df, aes(x = age.group, y = my.m))+geom_point() +
     geom_errorbar(aes(ymax = ci.upper, ymin = ci.lower),
                   width = 0.1)+
     xlab("Age Group")+
     ylab("Cholesterol level (mg/100ml)") +
  myTheme
```

Anything wrong? 

## Any interaction between Gender and Age group?
```{r}
GA.m <- with(combined.long.df, tapply(Cholesterol,
           list(Sex, Age.group), mean, na.rm = TRUE))
GA.m
```

## Calculating 95% CIs
```{r}
GA.sd <- with(combined.long.df,
              tapply(Cholesterol,
                     list(Sex, Age.group),
                          sd, na.rm = TRUE))

GA.n <- with(combined.long.df, tapply(Cholesterol,
      list(Sex, Age.group),
      function(x)length(which(!is.na(x)))))
GA.stder <- GA.sd/sqrt(GA.n)
GA.upper <- GA.m + 1.96*GA.stder
GA.lower <- GA.m - 1.96*GA.stder
```

## Calculating 95% CIs
```{r, warning=FALSE, fig.align='center'}
GA.df <- data.frame(
  Age.group = factor(rep(colnames(GA.m), 2),
                     levels = colnames(GA.m)),
  Gender = rep(rownames(GA.m), c(3, 3)),
  Mean = c(GA.m[1,], GA.m[2,]),
  Upper = c(GA.upper[1,], GA.upper[2,]),
  Lower = c(GA.lower[1,], GA.lower[2,])
  )
```

## Calculating 95% CIs
```{r, warning=FALSE, fig.align='center', echo = FALSE}
GA.df
```

## Plotting mean $\pm$ 95% CI:
```{r, warning=FALSE, fig.align='center', eval=FALSE}
g <- ggplot(GA.df, aes(x = Age.group, y = Mean,
                  color = Gender)) +
  xlab("Age Group")+
  ylab("Cholesterol level (mg/100ml)")

g + geom_point()+
  geom_errorbar(aes(ymax = Upper, ymin = Lower),
                width = 0.1) +
  myTheme

```

## Plotting mean $\pm$ 95% CI:
```{r, fig.height=3, fig.width=5, warning=FALSE, fig.align='center', echo = FALSE}
g <- ggplot(GA.df, aes(x = Age.group, y = Mean,
                  color = Gender)) +
  xlab("Age Group")+
  ylab("Cholesterol level (mg/100ml)")

g + geom_point() +
  geom_errorbar(aes(ymax = Upper, ymin = Lower),
                width = 0.1) +
  myTheme
```

## Side-by-side?
```{r, warning=FALSE, fig.align='center', eval=FALSE}
dodge <- position_dodge(width=0.2)

g + geom_point(position = dodge) +
  geom_errorbar(aes(ymax = Upper, ymin = Lower),
                width = 0.1, position = dodge) +
  myTheme
```

## Side-by-side
```{r, fig.height=3, fig.width=5, warning=FALSE, fig.align='center', echo = FALSE}
dodge <- position_dodge(width=0.2)

g + geom_point(position = dodge) +
  geom_errorbar(aes(ymax = Upper, ymin = Lower),
                width = 0.1, position = dodge) +
  myTheme
```

## Flip
```{r, fig.height=3, fig.width=5, warning=FALSE, fig.align='center', echo = FALSE}
dodge <- position_dodge(width=0.2)

g + geom_point(position = dodge) +
  geom_errorbar(aes(ymax = Upper, ymin = Lower),
                width = 0.1, position = dodge) + coord_flip() +
  myTheme
```

## Connect the points?
```{r, warning=FALSE, fig.align='center', eval=FALSE}
g + geom_point(position = dodge) +
  geom_errorbar(aes(ymax = Upper, ymin = Lower),
                width = 0.1, position = dodge) +
  geom_path(aes(x = as.numeric(Age.group)),
            position = dodge) +
  myTheme
```

## Connect the points
```{r, fig.height=3, fig.width=5, warning=FALSE, fig.align='center', echo = FALSE}
g + geom_point(position = dodge) +
  geom_errorbar(aes(ymax = Upper, ymin = Lower),
                width = 0.1, position = dodge) +
  geom_path(aes(x = as.numeric(Age.group)),
            position = dodge) +
  myTheme
```

##Save a ggplot
```{r, eval=FALSE}
ggsave("mtcars.pdf")
ggsave("mtcars.png")

ggsave("mtcars.pdf", width = 4, height = 4)
ggsave("mtcars.pdf", width = 20, height = 20, units = "cm")
```


##Summary

Plot Types     |  `geom` functions
-------------  |  -------------
Scatterplot   | `geom_point()`
Bars chart    | `geom_bar()`
Histogram     | `geom_histogram()`
Boxplot        | `geom_boxplot()`
Line plot     | `geom_path()`
Errorbar      | `geom_errorbar()`

- `ggplot2` Documentation:
[http://docs.ggplot2.org/current/](http://docs.ggplot2.org/current/)

- cheatsheets:
[https://www.rstudio.com/resources/cheatsheets/](https://www.rstudio.com/resources/cheatsheets/)




