---
title: "Introduction to $\\mathtt{R}$"
subtitle: "Session 2 -- Data subsetting"
author: "Statistical Consulting Centre"
date: "19 July, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = FALSE)
# knitr::opts_chunk$set(eval = FALSE)

lake.df <- read.csv("..\\..\\data\\Lake.csv", 
                      stringsAsFactors = FALSE)

```

# 1. Installing an \texttt{R} package
\texttt{R} packages are collections of user-defined functions. The function \texttt{std.error}, for example, is contained in the \texttt{plotrix} package.

1. Let's look at what happens when we try to use a function  before actually installing on our computer the package in which it is contained. E.g. Calculate the SEM of age using \texttt{std.error}.  

```{r, error = TRUE}
std.error(lake.df$pH)
```


2. Install the package \texttt{plotrix} while in your \texttt{R}  session by following the instructions below: 
  
    (a) Select \texttt{Packages} from the bottom right panel of your Rstudio interface.
    
    (b) Click on the \texttt{Install Packages} icon just below \texttt{Packages}.
    
    (c) Type \texttt{plotrix} in the blank space provided below
    ``\texttt{Packages (separate multiple with space or comma):}''
    
    (d) Click on \texttt{No} if you are asked you to restart \texttt{R}.
    
    (e) Submit the code \texttt{library(plotrix)} to the \texttt{R} console to make the functions contained in \texttt{plotrix} available in the current \texttt{R} session.

```{r}
library(plotrix)
```

3. Now, use \texttt{std.error} to calculate the standard error of the pH. 
```{r}
std.error(lake.df$pH)
```

4. Try writing your own code to calculate the standard error of the pH. Hint: This only requires one line of code. Use online resources if you cannot remember how the SEM is calculated.
```{r}
with(lake.df, sd(pH, na.rm = TRUE)/sqrt(length(pH)))
```


# 2. Write your own function
1.  In Session 2 you were shown a simple function to calculate the standard error of the mean (SEM), i.e.

```
  mystder <- function(x){
       mysd <- sd(x, na.rm = TRUE)
       n <- length(x)
       mysd/sqrt(n)
  }
```

 Type the above code into your \texttt{R} script and submit it to the \texttt{R} console. 

2. Modify the function in 2.1 so that the output will have only 2 decimal places.

```{r}
mystder <- function(x){
       mysd <- sd(x, na.rm = TRUE)
       n <- length(x)
       round(mysd/sqrt(n), 2)
}
```

3. Calculate the SEM of \texttt{pH} using the function you created in 2.2.
```{r}
mystder(lake.df$pH)
```


# 2. Subsetting datasets

1. Print the following to the console:
    
  - The pH of the first lake.
    
```{r}
lake.df$pH[1]
```

  - The pH of the last lake.
    
```{r}
lake.df$pH[53]
#or:
lake.df$pH[nrow(lake.df)]
```

  - The pH values of the first and last lakes.
    
```{r}
lake.df$pH[c(1, 53)]
```

  - All measurements made on the third lake.
    
```{r}
lake.df[3, ]
```

  - All pH values.
    
```{r}
lake.df[, "pH"]
```

2. Calculate:

  - The average pH of lakes with low Calcium concentrations.
    
```{r}
with(lake.df, mean(pH[Calcium == "Low"]))
```

  - The average pH of lakes with low Calcium concentrations and  Chlorophyll concentrations lower than 10.
    
```{r}
with(lake.df, mean(pH[Calcium == "Low" & Chlorophyll < 10]))
```

# 4. Challenge 

Modify the function given in 2., so that the function will return a 95\% confidence interval (with 2 decimal places). Hint: A 95\% confidence interval of a variable \texttt{x} is given by the mean of \texttt{x} $\pm$ 1.96 $\times$ SEM of \texttt{x}. You might find the \texttt{paste()} function useful.

```{r}
mystder <- function(x){
       mymean <- mean(x, na.rm = TRUE)
       mysd <- sd(x, na.rm = TRUE)
       n <- length(x)
       mystder = mysd/sqrt(n)
       upperCI = round(mymean + 1.96*mystder, 2)
       lowerCI = round(mymean - 1.96*mystder, 2)
       paste("(", lowerCI, " , ", upperCI, ")", sep = "")
}
mystder(lake.df$pH)
```
